### 2008-08-28 paper/paper.tex
Initial revision.
darcs-hash:20080828174936-bab43-6e3e1fb08b0c9c4272b9a9f4d4113b32c947484f.gz

\documentclass[runningheads]{llncs}

\usepackage{color}
\usepackage[colorlinks,pdftitle={dus},pdfauthor={James McKinna and Eelis van der Weegen}]{hyperref}

\spnewtheorem*{propo}{Proposition}{\bfseries}{\rmfamily}

%include polycode.fmt

%format pipe = "\mathopen{|}"

%format times = "\times"

%format . = "."
%format forall = "\forall"
%format Fixpoint = "\mathbf{Fixpoint}"
%format Definition = "\mathbf{Definition}"
%format Eval = "\mathbf{Eval}"
%format Notation = "\mathbf{Notation}"
%format Lemma = "\mathbf{Lemma}"
%format Proof = "\mathbf{Proof}"
%format Defined = "\mathbf{Defined}"
%format Function = "\mathbf{Function}"
%format Proposition = "\mathbf{Proposition}"
%format Record = "\mathbf{Record}"
%format Program = "\mathbf{Program}"
%format Inductive = "\mathbf{Inductive}"
%format Theorem = "\mathbf{Theorem}"
%format Let = "\mathbf{Let}"
%format Variable = "\mathbf{Variable}"
%format Variables = "\mathbf{Variables}"
%format measure = "\mathbf{measure}"

%format fcmp = "\circ"
%format match = "\mathbf{match}"
%format with = "\mathbf{with}"
%format end = "\mathbf{end}"
%format fun = "\lambda\hspace{-1mm} "
%format nat = "\mathbb{N}"
%format R = "\mathbb{R}"
%format exists = "\exists "
%format log2ceil = "\log_{2}\hspace{-0.5mm}"
%format over = "\mathbf{over}"
%format INR (a) = a
%format /\ = "\wedge "
%format \/ = "\vee"
%format MonoidMonadTrans.M = "\Varid{MMT}"
%format NatAddMonoid = (nat, 0, +)

%format sqrd (a) = (a)"^2"

%format `In` = "\in"
%format `NotIn` = "\notin"

%format C_MMT
%format ret_MMT
%format bind_MMT

%format NeTreeMonad = "M_\Varid{ne\_tree}"
%format M_ne_tree = "M_\Varid{ne\_tree}"
%format bind_ne_tree = "\Varid{bind}_\Varid{ne\_tree}"
%format ret_ne_tree = "\Varid{ret}_\Varid{ne\_tree}"
%format pick_ne_tree = "\Varid{pick}_\Varid{ne\_tree}"

%format Monoid_U
%format cmp_U
%format pick_U
%format partition_U
%format qs_U
%format U.qs = "\Varid{qs}_U"

%format M_NDP
%format le_NDP
%format qs_NDP

%format le_SP
%format ret_SP
%format bind_SP
%format qs_SP

% todo:
% - check that names are spelled and encoded right in bibliography
% - lhs2tex shows parameters declared with {...} differently from those declared with (...)
% - mention that we don't do correctness proofs (although we can add them in the formalization, of course)

\def\commentbegin{\quad\{\ }
\def\commentend{\}}

\def\typeset{}

\newcommand{\todo}[1]{\textcolor{magenta}{[TODO: #1]}}

\begin{document}

\nocite{*}

\title{A Machine-checked Proofs of Quicksort's Average-case Complexity, in Coq}

\authorrunning{\em}
\titlerunning{\em}

\author{James McKinna and Eelis van der Weegen}

\institute{Department of Computer Science, University of Nijmegen, the Netherlands}

\maketitle

\begin{abstract}
  We describe a machine-checked proof of Quicksort's $O(n \log n)$ average-case complexity, developed using the Coq system. To represent the algorithm, a shallow embedding is used in which the algorithm is defined as a monadically expressed functional program, with both the counted operation and the monad turned into parameters. Profiling instrumentation is then transparently inserted into this otherwise straightforward algorithm definition by instantiating it with the right monads. The instrumentation is sufficient to support formal yet convenient derivation of key properties of the algorithm's behavior needed by the proof---a step usually omitted in existing proofs. Thus, we directly take advantage of the computational nature of the type theory on which Coq is based.
\end{abstract}

\section{Introduction}

Proofs of Quicksort's $O(n \log n)$ average-case complexity are included in many textbooks on computational complexity \cite{introtoalgos} \todo{more}. This paper documents what the authors believe to be the first fully formal machine-checked version of such a proof, developed using the Coq proof assistant \cite{coq}.

The formalization is based on the ``paper proof'' in \cite{introtoalgos}, but adds a great deal of theory simply to handle all the details that are omitted in that original. In particular, the latter takes key assertions about the algorithm's behavior for granted. Indeed, it does not even formally define the algorithm in the first place. While this practice is common and perfectly reasonable for paper proofs intended for human consumption, it is a luxury we do not afford ourselves. Hence, a large part of this paper is devoted to the representations we use to define the algorithms, and how they support complexity proofs. These representations take the form of a shallow monadic embedding, and depend critically on the type theoretic nature of Coq's calculus \cite{coq'art}.

Section \ref{embed} introduces the basic idea of the embedding, and demonstrates its use by briefly proving Quicksort's quadratic worst-case complexity with it. In section \ref{nondetexpec}, the technique is expanded to handle the notion of the \emph{expected value} of non-deterministic programs, which then lets us state the main theorem in section \ref{statement}. Sections \ref{reduction} and \ref{finishing} detail the actual proof. Section \ref{conclusion} ends with conclusions and final remarks.

The Coq source files containing the entire formalization can be downloaded from \url{http://www.eelis.net/research/quicksort/}. The Coq version used is 8.2 beta 2.

\section{A Shallow Monadic Embedding}
\label{embed}

Before the main theorem can even be formally stated, let alone proved, a formal definition of the algorithm is needed, expressed in a programming language with well-defined semantics, suitable for subsequent formal reasoning. Here, we exploit the computational nature of the type theory on which Coq is based. Rather than defining the syntax and semantics of a programming language from scratch (this would be a ``deep'' embedding) like we would have to do in, say, ZFC mathematics, we instead use the built-in lambda calculus as a functional programming language, and write the algorithm as a specially adorned lambda term. More specifically, we express the algorithm monadically, with both the monad and the counted operation (in our case, the elementwise comparison---the usual complexity metric for sorting algorithms) made into parameters. Expressed this way, a deterministic quicksort that simply selects the head of the input list as its pivot element, and uses two simple filter passes to partition the input list, looks as follows:

\begin{code}
  Variables (M: Monad) (T: Set) (le: T -> T -> M bool).

  Definition gt (x y: T): M bool := liftM negb (le x y).

  Fixpoint filter (c: T -> M bool) (l: list T): M (list T) :=
    match l with
    | nil => ret nil
    | h :: t =>
      t' <- filter c t;
      b <- c h;
      ret (if b then h :: t' else t')
    end.

  Program Fixpoint qs (l: list T) {measure length l}: M (list T) :=
    match l with
    | nil => ret nil
    | pivot :: t =>
        lower <- filter (gt pivot) t >>= qs;
        upper <- filter (le pivot) t >>= qs;
        ret (lower ++ pivot :: upper)
    end.
\end{code}
|Monad| is a dependent record containing the (coercible) carrier of type |Set -> Set|, along with |bind| (infix: |>>=|) and |ret| (for ``return'') operations, and the three monad laws. For a general introduction to monadic programming and monad laws, see \cite{wadler93monads}. The notation |x <- y; z|, shorthand for |y >>= \x -> z| (where |x| may occur in |z|) is known as ``do-notation''.

We use Coq's |Program| facility \cite{subsetcoercions} to cope with the fact that Quicksort's recursion is non-structural, by using the input list length as a measure. The proof obligations generated by |Program| for the above definition and measure are trivial enough for Coq to prove mostly by itself.

By instantiating the above definitions with the right monad, we can transparently insert comparison-counting instrumentation into the algorithm, which will prove to be sufficient to let us reason about its complexity. But before we do so, let us note that if the above definitions are instead instantiated with the identity monad and an ordinary elementwise comparison on |T|, then the monadic scaffolding melts away, and the result is equivalent to an ordinary non-instrumented, non-monadic version, suitable for extraction and correctness proofs. This means that while we will instantiate the definitions with less trivial monads to support our complexity proofs, we can take some comfort in knowing that the object of those proofs is, in a very concrete sense, the actual Quicksort algorithm (as one would write it in a functional programming language), rather than some idealized model thereof.

For reasons that will become clear in later sections, we compose the monad we will instantiate the above definitions with using a monad transformer called |MMT| (for ``monoid monad transformer''), which piggybacks a monoid onto an existing monad.

\begin{code}
  Variables (monoid: Monoid) (monad: Monad).

  Let C_MMT: Set -> Set := monad fcmp prod monoid.

  Let ret_MMT (T: Set): T -> C_MMT T := ret fcmp pair (monoid_zero monoid).

  Let bind_MMT (A B: Set) (a: C_MMT A) (ab: A -> C_MMT B): C_MMT B :=
    x <- a; y <- ab (snd x); ret (monoid_mult monoid (fst x) (fst y), snd y).

  Definition MMT: Monad := Build_Monad C_MMT bind_MMT ret_MMT.
\end{code}
(In the interest of brevity, we omit proofs of the monad laws for |MMT| and all other monads defined in this paper. These proofs can all be found in the Coq code.)

We use |MMT| to piggyback the additive monoid on |nat| onto the identity monad, and lift elementwise comparison into the resulting monad, which we call |SP| (for ``simply-profiled'').
\begin{code}
  Definition SP: Monad := MMT NatAddMonoid IdMonad.

  Definition le_SP (x y: nat): SP bool := (1, le x y).
\end{code}
When instantiated with this monad and comparison operation, |qs| produces the comparison count as part of its result.
\begin{code}
  Definition qs_SP := qs SP le_SP.

  Eval compute in qs_SP (3 :: 1 :: 0 :: 4 :: 5 :: 2 :: nil).
    = (16, 0 :: 1 :: 2 :: 3 :: 4 :: 5 :: nil)
\end{code}
Defining |cost| and |result| as the first and second projection, respectively, we have the identities
\begin{code}
  forall x, cost (ret_SP x) = 0,
  forall x f, cost (bind_SP x f) = cost x + cost (f (result x),
  forall x y, cost (le_SP x y) = 1.
\end{code}
This very modest amount of machinery is sufficient for a straightforward proof of Quicksort's quadratic worst-case complexity.

\begin{propo}|qs_worst: forall l, cost (qs_SP l) <= sqrd (length l).|\end{propo}
  \todo{get rid of period after ``Proposition'' above}
\begin{proof}
  The proof is by induction matching |qs|'s recursion. For an empty input list, we have |cost (qs_SP nil) = cost (ret nil) = 0 <= sqrd (length l)|. For a non-empty input list |(pivot :: t)|, the cost decomposes into
  \begin{code}
    cost (filter (le pivot) t) + cost (qs_SP (result (filter (lt pivot) t))) +
    cost (filter (gt pivot) t) + cost (qs_SP (result (filter (gt pivot) t))) +
    cost (ret (result (qs_SP (result (filter (lt pivot) t))) ++
     pivot :: result (qs_SP (result (filter (gt pivot) t))))).
  \end{code}
  The |filter| costs are easily proved (by induction on |t|) to be |length t| each. The cost of the final |ret| is 0 by definition. The induction hypothesis applies to the recursive |qs_SP| calls. Furthermore,
  \begin{code}
  length (result (filter (le pivot) t)) +
  length (result (filter (gt pivot) t)) <= length t
  \end{code}
  can easily be proved by induction on |t|, because the two predicates filtered on are mutually exclusive. Abstracting |filter (le pivot) t| and its |gt| counterpart, this leaves
  \begin{code}
    forall (t flt flt': list T), length flt + length flt' <= length t ->
      length t + sqrd (length flt) + length t + sqrd (length flt')  + 0 <= sqrd (S (length t)),
  \end{code}
  which is true by elementary arithmetic. \qed
\end{proof}

We now expand the machinery in preparation of the average-case complexity proof.

\section{Nondeterminism and Expected Values}
\label{nondetexpec}

The Quicksort algorithm used by the average-case proof we formalize differs from the one presented in the last section in two ways: it is nondeterministic, and uses a single three-way partition pass instead of two two-way filter passes. Combined, these two traits ensure that the $O(n \log n)$ average-case property holds not just averaged over all input lists, but for each individual input list as well; nondeterministic pivot selection avoids the pathological cases any deterministic pivot selection strategy inevitably suffers, while a single three-way partition pass avoids the pathological quadratic case that two two-way filter passes suffer for an input list consisting of elements all of the same value. This greatly simplifies things, because it means that the global bound will follow immediately if it is proved for an arbitrary input.

Hence, we define

\parbox{\textwidth}{\begin{code}
  Variables (T: Set) (M: Monad) (cmp: T -> T -> M comparison).

  Fixpoint partition (pivot: T) (l: list T): M (Partitioning T) :=
    match l with
    | nil => ret emptyPartitioning
    | h :: t =>
        b <- cmp h pivot;
        tt <- partition pivot t ;
        ret (addToPartitioning b h tt)
    end.

  Variable pick: forall (A: Set), ne_list A -> M A.

  Program Fixpoint qs (l: list T) {measure length l}: M (list T) :=
    match l with
    | nil => ret nil
    | h :: t =>
        i <- pick [0 .. length t];
        let pivot := nth (h :: t) i in
        part <- partition pivot (remove (h :: t) i);
        low <- qs (part Lt);
        upp <- qs (part Gt);
        ret (low ++ pivot :: part Eq ++ upp)
    end.
\end{code}}
Here, |comparison| is an enumeration with values |Lt|, |Eq|, and |Gt|. A |Partitioning T| is a function of type |comparison -> list T|. An |ne_list T| is a non-empty list of |T|'s. The functions |nth| and |remove| select and remove the $n$th element of a list, respectively.

Nondeterminism can now be achieved by instantiating these definitions with a suitable monad, along with a corresponding |pick| operation. A deterministic, non-instrumented version can still be obtained, simply by instantiating the definitions with the identity monad and a corresponding deterministic |pick| operation, like |head|.

Let us now consider what kind of nondeterminism monad would be suitable for reasoning about the expected value of a nondeterministic program. The list monad is commonly used to emulate nondeterministic computation. With the list monad, the program
\begin{code}
  x <- pick [0, 1]; if x = 0 then ret 0 else pick [1, 2],
\end{code}
produces |[0, 1, 2]| as its list of possible outcomes. Unfortunately, the information that 0 is a more likely outcome than 1 or 2, has been lost. Such relative probabilities are critical to the notion of an expected value: the expected value of the program above is |avg [0, avg [1, 2]] = {-"\frac{3}{4}"-} /= 1 = avg [0, 1, 2]|. This makes list nondeterminism unsuitable for our purposes.

Using tree nondeterminism instead solves the problem.
\begin{code}
  Inductive ne_tree (T: Set): Set :=
    | Leaf: T -> ne_tree T
    | Node: ne_list (ne_tree T) -> ne_tree T.

  Definition ret_ne_tree {A: Set}: A -> C A := Leaf.

  Fixpoint bind_ne_tree (A B: Set) (m: C A) (k: A -> C B): C B :=
    match m with
    | Leaf a => k a
    | Node ts => Node (ne_list.map (fun x => bind_ne_tree x k) ts)
    end.

  Definition M_ne_tree: Monad := Build_Monad ne_tree bind_ne_tree ret_ne_tree.

  Definition pick_ne_tree (T: Set): ne_list T -> M_ne_tree T
    := Node fcmp ne_list.map Leaf.
\end{code}
(We use non-empty trees because are not interested in computations that produce no values at all, and using potentially empty trees would complicate the definition of a tree's average value below.)

With this monad and pick operation, the same program now produces the tree |Node [Leaf 0, Node [Leaf 1, Leaf 2]]|, which preserves the relative probabilities. The expected value now coincides with the weighted average of these trees:
\begin{code}
  Definition ne_tree.avg: ne_tree R -> R := ne_tree.fold id ne_list.avg.
\end{code}

Relative probabilities are also the reason we use an $n$-ary choice primitive rather than a binary one, because correctly emulating (that is, without skewing the relative probabilities) an $n$-ary choice by a sequence of binary choices is only possible when $n$ is a power of two.

Because we are often interested in the expected value of a measure of the output of a program that does not return numbers, we further define

\begin{code}
  Definition expec (T: Set) (f: T -> nat): ne_tree T -> R
    := ne_tree.avg fcmp ne_tree.map f.
\end{code}

Thus, given a program |P| of type |M_ne_tree (list bool)|, |expec length P| denotes the expected length of the result list, if we interpret values of type |M_ne_tree T| as nondeterministically computed values of type |T|.

To form the monad with which we will instantiate |qs| for the main theorem, we now piggyback the additive monoid on |nat| onto |M_ne_tree| using |MMT|, and call the result |NDP| (for ``non-deterministically profiled''):
\begin{code}
  Definition M_NDP: Monad := MMT NatAddMonoid M_ne_tree.

  Definition le_NDP (x y: T): M bool := ret_ne_tree (1, le x y).

  Definition qs_NDP := qs M_NDP le_NDP (lift pick_ne_tree).
\end{code}

When working with computations in a monad formed by transforming |M_ne_tree| using |MMT|, we are often interested in the expected value of a function of the piggybacked monoid. For this, we introduce |monoid_expec|.
\begin{code}
  Definition monoid_expec (m: Monoid) (f: m -> nat) {A: Set}
    : MonoidMonadTrans.M m NeTreeMonad A -> R := expec (f fcmp fst).
\end{code}
Thus, we have |forall t, expec cost t = monoid_expec id t|.

Like |expec|, |monoid_expec| gives rise to several identities. One identity that we would like to highlight is
\begin{code}
  monoid_expec_plus: forall (m: Monoid) (h: m -> nat) (A B: Set)
    (f: MMT m NeTreeMonad A) (g: A -> MMT m NeTreeMonad B):

    (forall x y `In` f -> monoid_expec h (g (snd x)) = monoid_expec h (g (snd y))),
    monoid_expec h (f >>= g) =
      monoid_expec h f + monoid_expec h (g (snd (ne_tree.head f))),
\end{code}
which states that if one transforms |M_ne_tree| using a monoid |m|, then for a monoid homomorphism |h| from |m| to the additive monoid on |nat|, |monoid_expec h| distributes over |bind|, provided that the expected monoid value of the right hand side does not depend on the computed value of the left hand side. Since |id| is a monoid homomorphism, |monoid_expec_plus| applies to |NDP| and |expec cost|. In section \ref{reduction}, we will use it with another monoid and homomorphism.

\section{The Statement}
\label{statement}

Now that we have a solid definition of nondeterministic, instrumented quicksort, as well as the means to express the expected values of nondeterministic programs, the last thing needed before the main theorem can be stated, is the notion of big-O complexity. The definition we use is a standard textbook definition, except that we work with a separate measure function parameter:

\parbox{\textwidth}{\begin{code}
  Definition measured_bigO (X: Set)
    (m: X -> nat) (f: X -> R) (g: nat -> R): Prop
      := exists c, exists n, forall x, n <= m x -> f x <= c * g (m x).

  Notation {-"``"-}over m ,{-"\ "-} f =O( g ){-"\text{''''}"-} := (measured_bigO m f g).
\end{code}}
We now state the main theorem.
\begin{code}
  Theorem qs_avg: over length,
    expec cost fcmp qs_NDP =O(fun n => INR (n * log2ceil n)).
\end{code}
Thanks to the property discussed at the start of the previous section, the above follows as a corollary from the (slightly stronger) statement that
\begin{code}
  Theorem qs_expec_cost:
    forall l, expec cost (qs_NDP l) <= 2 * length l * S (log2ceil (length l)),
\end{code}
the proof of which is described in the remainder of this paper.

\section{Reduction to Pairwise Comparison Counts}
\label{reduction}

The main idea in the proof is to reduce |qs_expec_cost| to a statement about the expected number of comparisons between specific pairs of elements from the input list.

If $X \equiv X_{I_0} \ldots X_{I_{n-1}}$ is the input list, with $I$ a permutation of |[0 ... n-1]| such that $X_0 \ldots X_n$ is sorted, then the expected number of comparisons between any $X_i$ and $X_j$ with $i < j$ is at most $2 / S (j - i)$. In other words, the expected number of comparisons between two list elements is bounded by a simple function of the number of list elements that separate the two in the sort order. We prove this fact in the next section, but first show how |qs_expec_cost| follows from it.

Combined with the observation that the total expected number of comparisons equals the sum of the expected numbers of comparisons for each |(i, j)| in $\Varid{IJ} := \{ (i, j) \in [0, |length l|) \mid i < j \}$, the property described above suggests breaking up the inequality into
\begin{code}
  expec cost (qs_NDP l) <= {-"\displaystyle{\sum_{(i, j) \in \Varid{IJ}}\frac{2}{S (j - i)}}"-} <= 2 * length l * S (log2ceil (length l)).
\end{code}
Proving the right inequality requires a bit of analysis involving the harmonic series. This part of the proof could be fairly directly transcribed from the paper proof, using the existing real-number theory in the Coq standard library, with few complications and additions. We refer the interested reader to the paper proof.

The left inequality is the challenging one. To bring it closer to the index summation, we first rewrite
\begin{code}
  expec cost (qs_NDP l) = expec cost (qs_NDP (map (nth (sort l)) li)),
\end{code}
where |sort| may be any sorting function (including |qs| itself), and where |li| is a permutation of |[0...n - 1]| such that |map (nth (sort l)) li = l| (such an |li| can easily be proven to exist).

Next, we introduce a specialized monad and comparison operation that go one step further in focusing specifically on these indices.
\begin{code}
  Definition Monoid_U: Monoid := (nat * nat, nil, ++).

  Definition U: Monad := MonoidMonadTrans.M Monoid_U NeTreeMonad.

  Definition lookup_cmp (x y: nat): comparison :=
    cmp (nth (sort l) x) (nth (sort l) y).

  Definition cmp_U (x y: nat): U comparison :=
    ret ((if x <= y then (x, y) else (y, x)) :: nil, lookup_cmp x y).

  Definition qs_U: list nat -> list nat := qs cmp_U pick_U.
\end{code}
|qs_U| operates directly on a list of natural numbers representing indices into |sort l|. Comparison of indices is defined by comparison of the |T| values they denote in |sort l|. Furthermore, rather than producing a grand total comparison count the way |NDP| does, |U| records every pair of compared indices, by using |MMT| with the free monoid over |nat * nat| pairs, instead of the additive monoid on |nat| we used up until now.

The goal can now be rewritten using
\begin{code}expec cost (qs_NDP (map (nth (sort l)) li))
  = monoid_expec length (qs_U li) = expec (length fcmp fst) (qs_U li).
\end{code}
The first equality is justified by a separate induction matching either side's |qs|'s recursion, proving that counting comparisons as they appear is really the same as recording them in a list and then computing the length of the resulting list.

After rewriting once more, this time using\begin{code}expec_fcmp: forall f g t, expec (f fcmp g) t = expec f (ne_tree.map g t),\end{code}the goal becomes
\begin{code}
  expec length (ne_tree.map fst (qs_U li)) <= {-"\displaystyle{\sum_{(i, j) \in \Varid{IJ}}\frac{2}{S (j - i)}}"-}.
\end{code}
We now invoke another lemma which bounds a nondeterministically computed list's expected length by the expected number of occurrences of specific values in that list. More specifically, it states that

\parbox{\textwidth}{\begin{code}
  forall (X: Set) (fr: X -> R) (q: list X) (t: ne_tree (list X)):
    (forall x `In` q, expec (count x) t <= fr i) ->
    (forall x `NotIn` q, expec (count x) t = 0) ->
    NoDuplicates q -> expec length t <= {-"\displaystyle{\sum_{\Varid{i} \in \Varid{q}}\Varid{fr}\ \Varid{i}}"-}.
\end{code}}
The last of the three subgoals thus generated states that there are no duplicate values in |IJ|, which trivially holds. The second subgoal generated is\begin{code}forall (i, j) `NotIn` IJ, expec (count (i, j)) (ne_tree.map fst (qs_U li)) = 0.\end{code} Rewriting this using |expec_fcmp| backwards, then rewriting the |expec| as a |monoid_expec|, and then generalizing the premise, results in the following property which is useful in its own right, as we will see later.
\begin{code}sound_cmp_expec_0: forall i j li, (i `NotIn` li \/ j `NotIn` li) ->
  monoid_expec (count (i, j)) (U.qs li) = 0,\end{code}
The proof of |sound_cmp_expec_0|, which we won't show here, is by a separate induction matching |qs|'s recursion.

This leaves but one subgoal which, expressed using a |monoid_expec|, reads
\begin{code}
  forall (i, j) `In` IJ, monoid_expec (count (i, j)) (qs_U li) <= 2 / INR (S (j - i)).
\end{code}
This is exactly  to the property described at the beginning of this section. We prove it in the next section.

\section{Finishing the Proof}
\label{finishing}

To get a stronger induction hypothesis, we slightly generalize the goal to
\begin{code}
  forall i j, i < j -> forall (li: list nat) (b: nat), IndexSeq b li ->
    monoid_expec (count (i, j)) (qs_U li) <= 2 / INR (S (j - i)).
\end{code}
The |(i, j) `In` IJ| hypothesis is dropped, because the statement is also true if |(i, j) `NotIn` IJ|, per |sound_cmp_expec_0|. The |IndexSeq| premise expresses that |li| is a permutation of |[b...b+length l]| (for some |b|).

Again, the proof is by induction matching |qs|'s recursion. In the base case, |li| is |nil|, and the left side of the inequality reduces to 0. In the recursive case, |qs| unfolds:
\begin{code}
monoid_expec (count (i, j)) (
    pi <- pick [0 ... n];
    let pivot := nth li pi in
    part <- partition_U pivot (remove li pi);
    low <- qs_U (part Lt);
    upp <- qs_U (part Gt);
    ret (low ++ pivot :: part Eq ++ upp)
  ) <= 2 / INR (S (j - i))
\end{code}
Since |cmp_U| is deterministic, |partition_U| is deterministic here as well. Furthermore, since we know exactly what monadic effects |partition| has in this case, we can split those effects off and then revert to simple uneffectful |filter| passes. Finally, the outer |monoid_expec| can be partially evaluated because of the immediately visible |pick|. Using these observations, the goal can be rewritten into a form that uses less monadic indirection:

\begin{code}
avg (map (monoid_expec (count (i, j)) fcmp (fun pi =>
    let pivot := nth li pi in
    let rest := remove li pi in
    ne_tree.map (map_fst (++ map (U.unordered_nat_pair pivot) rest)) (
      lower <- qs_U (filter ((= Lt) fcmp lookup_cmp pivot) rest);
      upper <- qs_U (filter ((= Gt) fcmp lookup_cmp pivot) rest);
      ret (lower ++ (pivot :: filter ((= Eq) fcmp lookup_cmp pivot) rest) ++ upper)
    ))) [0...n]) <= 2 / INR (S (j - i)).
\end{code}
Here, |map_fst| applies a function to a pair's first component.
% todo: get this [0...n]/[0...n-1] stuff right.

We now distinguish between five different cases that can occur for the nondeterministically picked |pivot|; it can either be less than |i|, equal to |i|, between |i| and |j|, equal to |j|, or greater than |j|. Each case occurs a certain number of times, and has an associated expected number of |(i, j)| comparisons (coming either from the |map_fst| term representing the |partition| pass, or from the two recursive |qs_U| calls). To represent this split, we first rewrite the right side of the inequality to\begin{code}(2 / S (j - i) * (i - b) + 1 + 0 + 1 + 2 / S (j - i) * (b + n - j)) / S n.\end{code}
This form reflects the facts that
\begin{itemize}
\item the case where |pivot| is less than |i| occurs |i - b| times, and in each instance, the expected number of |(i, j)| comparisons is no more than |2 / S (j - i)|;
\item the case where the |pivot| is equal to |i| occurs once, and in this case no more than a single |(i, j)| comparison is expected;
\item in the case where |pivot| lies between |i| and |j|, the number of expected |(i, j)| comparisons is 0, and consequently it does not matter how many times this case occurs;
\item the case where the |pivot| is equal to |j| occurs once, and in this case no more than a single |(i, j)| comparison is expected;
\item the case where the |pivot| is greater than |j| occurs |b + n - j| times, and in each instance, the expected number of |(i, j)| comparisons is no more than |2 / S (j - i)|.
\end{itemize}
With the right side of the inequality in this form, we unfold the |avg| application on the left into |sum (...) / S n|, and then cancel the division by |S n| on both sides. Next, to actually realize the split, we apply a specialized lemma stating that
\begin{code}
   forall b i j X f n (li: list nat)
     (g: [0..n] -> U X), IndexSeq b li ->
    b <= i < j < b + S n -> forall ca cb, 0 <= ca -> 0 <= cb ->
    (forall pi, nth li pi < i -> expec f (g pi) <= ca) ->
    (forall pi, nth li pi = i -> expec f (g pi) <= cb) ->
    (forall pi, i < nth li pi < j -> expec f (g pi) = 0) ->
    (forall pi, nth li pi = j -> expec f (g pi) <= cb) ->
    (forall pi, j < nth li pi -> expec f (g pi) <= ca) ->
      sum (map (expec f fcmp g) [0..n]) <=
        ca * (i - b) + cb + 0 + cb + ca * (b + n - j).
\end{code}
The proof of this lemma is somewhat tedious \todo{because..}.
% involving a lot of repetitive arguments about permutations, etc

Five subgoals remain after applying the above lemma---one for each of the listed cases. The first one reads
\begin{code}
  forall pi,
    let pivot := nth li pi in
    let rest := remove li pi in
      pivot < i ->
      monoid_expec (count (i, j))
        (ne_tree.map (map_fst (++ map (U.unordered_nat_pair pivot) rest)) (
          foo <- qs_U (filter ((= Lt) fcmp lookup_cmp pivot) rest);
          bar <- qs_U (filter ((= Gt) fcmp lookup_cmp pivot) rest);
          ret (foo ++ (pivot :: filter ((= Gt) fcmp lookup_cmp pivot) rest) ++ bar)))
        <= 2 / S (j - i).
  \end{code}
Rewriting using a lemma saying that
\begin{code}
  forall (A: Set) (g: m) (h: m -> nat) (t: MMT m NeTreeMonad A):
    MonoidHomomorphism h ->
    monoid_expec h (ne_tree.map (map_fst (monoid_mult m g)) t) =
    INR (h g) + monoid_expec t,
\end{code}
the head reduces to
\begin{code}
    count (i, j) (map (U.unordered_nat_pair pivot) rest) +
    monoid_expec (count (i, j))
      (foo <- qs_U (filter ((= Lt) fcmp lookup_cmp pivot) rest);
      bar <- qs_U (filter ((= Gt) fcmp lookup_cmp pivot) rest);
      ret  (foo ++ (nth v pi :: filter ((= Eq) fcmp lookup_cmp pivot) rest) ++ bar))
    <= 2 / S (j - i).
\end{code}
From |pivot < i| and |i < j|, we have |pivot < j|. Since each of the comparisons in |map (U.unordered_nat_pair pivot) rest| involves the pivot element, it follows that none of them can represent comparisons between |i| and |j|. Hence, the first term vanishes. Furthermore, since |count (i, j)| is a monoid homomorphism, |monoid_expec_plus| lets us distribute |monoid_expec|. Since the |ret| term does not produce any comparisons either (by definition), its |monoid_expec| term vanishes, too. What remains are the two recursive calls:

\begin{code}
monoid_expec (count (i, j))
  (qs_U (filter ((= Lt) fcmp lookup_cmp pivot) rest)) +
monoid_expec (count (i, j))
  (qs_U (filter ((= Gt) fcmp lookup_cmp pivot) rest)) <= 2 / INR (S (j - i))
\end{code}
All indices in the lower filtered part denote elements that compare less than the element denoted by the pivot. Since the former precede the latter in |sort l|, it must be the case that the indices areless than |pivot|. By |sound_cmp_expec_0|, it follows that the first |qs_U| term will produce no |(i, j)| comparisons, so the first |monoid_expec| term vanishes as well, leaving
\begin{code}
monoid_expec (count (i, j))
  (qs_U (filter ((= Gt) fcmp lookup_cmp pivot) rest)) <= 2 / INR (S (j - i)).
\end{code}
We now compare |nth (sort l) i| with |nth (sort l) pivot|.
\begin{itemize}
\item If the two are equal, then |i| will not occur in the |filter term|, allowing us to invoke |sound_cmp_expec_0| yet again.
\item If |nth (sort l) i < nth (sort l) pivot|, then we must have |i < pivot|, contradicting the assumption that |pivot < i|.
\item If |nth (sort l) i > nth (sort l) pivot|, then we apply the induction hypothesis. For this, it must be shown that the filter term is a proper contiguous index sequence, which we relegated to a lemma we won't show here.
\end{itemize}
This concludes the case where |i < pivot|. The case where |j < pivot| is symmetric. The other cases use similar arguments.

The proof is now complete.

\section{Final Remarks}
\label{conclusion}

In the interest of brevity, we have omitted innumerable details and lemmas in the description of the proof. Still, the parts shown are reasonably faithful to the actual formalization. For completeness, we briefly list some complicating factors that we have ignored. In some cases, resolving these took significant effort.

\begin{itemize}
\item We have pretended to have used ordinary natural numbers as indices into ordinary lists, completely ignoring the inescapable issues of index validity that could not be ignored in the actual formalization. There, we use vectors (lists whose size is part of their type) and bounded natural numbers in many places instead. Using these substantially reduces the amount of |i < length l| proofs that need to be produced, converted, and passed around, but the issue is still far from painless.
\item Using the |Program| facility to deal with quicksort's non-structural recursion was not completely as trivial as we made it out to be. To make proving the generated proof obligations possible, the types of |filter| and |partition| had to be sigma-decorated with modest length guarantees, because the opacity of the unspecified monad's |bind| operation prevented |Program| from being able to ``see'' the relation between the function's parameters and the arguments it passed in its recursive calls.
\item The definitions produced by |Program| use clever recursion operators that translate a well-founded relation into a form that can be structurally recursed on. Consequently, corresponding induction principles are similarly clever. Using these induction principles for a function already defined with two layers of monadicity quickly became unwieldy. To deal with this, we defined a series of induction principles specialized for specific instances of |qs|, which put the recursive proof obligation in as simple terms as possible, with as little monadic indirection as possible, using among other things the observations regarding determinism of the comparison operation, as described in section \ref{finishing}.
\end{itemize}

For all the gruesome details, we refer the interested reader to the Coq source files.

In conclusion, we have found the shallow monadic embedding to be a straightforward, clean, and solid basis for reasoning about an algorithm's computational complexity. Calling it an embedding almost seems misplaced, since the algorithm definitions are essentially just monadically expressed functional programs in Coq's calculus. Furthermore, since monads and monad transformers have been used in functional programming for decades, the basic idea should be immediately clear to any functional programmer. Still, it is worth noting that this pleasing convergence of subject and object languages is due solely to the fact that Coq is based on type theory.

Quicksort turned out to be a good stress test for the technique, showing that it copes well with nondeterminism and nontrivial average-case bounds. The |U| monad shows that custom profiling monads can be defined to analyse very algorithm-specific traits for a specific part of a larger complexity proof, which was crucial for this particular proof.

For future work, a logical next challenge is to try and formalize a (likely more complex) proof that the $O(n \log n)$ bound holds also for deterministic quicksorts (with sensible pivot selection strategies).

\bibliographystyle{plain}
\addcontentsline{toc}{chapter}{Bibliography}
\bibliography{paper}

\end{document}

### 2008-08-28 paper/compile.sh

@@ -1,0 +1,12 @@
+#!/bin/sh
+
+lhs2TeX < paper.tex > processed.tex
+
+set -e
+pdflatex processed.tex
+bibtex processed
+pdflatex processed.tex
+bibtex processed
+pdflatex processed.tex
+cat processed.pdf > paper.pdf
+rm processed.*

### 2008-08-29 paper/paper.tex
paper: Fix title.
darcs-hash:20080828175112-bab43-debeb3343db819e77f00903149d02b72c8d6a3a2.gz
paper: New abstract.
darcs-hash:20080829013010-bab43-f77c78689535d9761c1b7a1857f519aefaa70670.gz

@@
 
-\title{A Machine-checked Proofs of Quicksort's Average-case Complexity, in Coq}
+\title{A Machine-checked Proof of the Average-case Complexity of Quicksort in Coq}
 
@@
 \begin{abstract}
-  We describe a machine-checked proof of Quicksort's $O(n \log n)$ average-case complexity, developed using the Coq system.
-To represent the algorithm, a shallow embedding is used in which the algorithm is defined as a monadically expressed functional program, with both the counted operation and the monad turned into parameters.
-Profiling instrumentation is then transparently inserted into this otherwise straightforward algorithm definition by instantiating it with the right monads.
-The instrumentation is sufficient to support formal yet convenient derivation of key properties of the algorithm's behavior needed by the proof---a step usually omitted in existing proofs.
-Thus, we directly take advantage of the computational nature of the type theory on which Coq is based.
+As a case-study in machine-checked reasoning about the complexity of algorithms in type theory, we describe a proof of the average-case complexity of Quicksort in Coq.
+The proof attempts to follow a textbook development, at the heart of which lies a technical lemma about the behavior of the algorithm for which the original proof only gives an intuitive justification.
+Our formalization includes a fully formal treatment of this lemma, introducing a number of technical innovations:
+a shallow embedding of the algorithm as a monadically expressed functional program, instantiated with a variety of operation-counting monads that capture worst- and average case complexity of deterministic and non-deterministic programs, generalized to count in an arbitrary monoid (to support a key step in the proof where the expected comparison count is translated into the expected length of a recorded list of all comparisons);
+a small theory of expectation for non-deterministically computed values, featuring both general map-fusion like results, and specific counting arguments.
 \end{abstract}

### 2008-08-29 paper/paper.tex
paper: Various minor fixes re. meeting 2008-08-28.
darcs-hash:20080829023655-bab43-452101cfa5cb230245a37024a0421df5939a84fc.gz

@@
 
-  Let C_MMT: Set -> Set := monad fcmp prod monoid.
+  Let C_MMT (T: Set): Set := monad (monoid * T).
 
@@
     | nil => ret nil
-    | h :: t =>
-        i <- pick [0 .. length t];
-        let pivot := nth (h :: t) i in
-        part <- partition pivot (remove (h :: t) i);
+    | _ =>
+        i <- pick [0 ... length l - 1];
+        let pivot := nth l i in
+        part <- partition pivot (remove l i);
         low <- qs (part Lt);
@@
 
-  Definition ret_ne_tree {A: Set}: A -> C A := Leaf.
+  Definition ret_ne_tree {A: Set}: A -> ne_tree A := Leaf.
 
-  Fixpoint bind_ne_tree (A B: Set) (m: C A) (k: A -> C B): C B :=
+  Fixpoint bind_ne_tree (A B: Set) (m: ne_tree A) (k: A -> ne_tree B): ne_tree B :=
     match m with
@@
 \end{code}
-(We use non-empty trees because are not interested in computations that produce no values at all, and using potentially empty trees would complicate the definition of a tree's average value below.)
+(We use non-empty trees because we do not consider computations that produce no values at all, and using potentially empty trees would complicate the definition of a tree's average value below.)
 
@@
   Definition monoid_expec (m: Monoid) (f: m -> nat) {A: Set}
-    : MonoidMonadTrans.M m NeTreeMonad A -> R := expec (f fcmp fst).
+    : (MonoidMonadTrans.M m NeTreeMonad A) -> R := expec (f fcmp fst).
 \end{code}
-Thus, we have |forall t, expec cost t = monoid_expec id t|.
+Since |cost = fst|, we have |forall t, expec cost t = monoid_expec id t|.
 
@@
     (f: MMT m NeTreeMonad A) (g: A -> MMT m NeTreeMonad B):
-
     (forall x y `In` f -> monoid_expec h (g (snd x)) = monoid_expec h (g (snd y))),
@@
 \begin{code}
-  Definition Monoid_U: Monoid := (nat * nat, nil, ++).
+  Definition Monoid_U: Monoid := (list (nat * nat), nil, ++).
 
@@
 Comparison of indices is defined by comparison of the |T| values they denote in |sort l|.
-Furthermore, rather than producing a grand total comparison count the way |NDP| does, |U| records every pair of compared indices, by using |MMT| with the free monoid over |nat * nat| pairs, instead of the additive monoid on |nat| we used up until now.
+Furthermore, rather than producing a grand total comparison count the way |NDP| does, |U| records every pair of compared indices, by using |MMT| with |Monoid_U|, the free monoid over |nat * nat| pairs, instead of the additive monoid on |nat| we used up until now.
 
@@
 \end{code}
-The first equality is justified by a separate induction matching either side's |qs|'s recursion, proving that counting comparisons as they appear is really the same as recording them in a list and then computing the length of the resulting list.
+The first equality expresses that the expected number of comparisons counted by |NDP| is equal to the expected length of the list of comparisons recorded by |U|.
+In the code, it is proved as a separate lemma by induction matching either side's |qs|'s recursion.
+The second equality merely unfolds the definition of |monoid_expec|.
 
@@
 
-This leaves but one subgoal which, expressed using a |monoid_expec|, reads
+This leaves but one subgoal which, expressed with |monoid_expec|, reads
 \begin{code}
@@
 All indices in the lower filtered part denote elements that compare less than the element denoted by the pivot.
-Since the former precede the latter in |sort l|, it must be the case that the indices areless than |pivot|.
+Since the former precede the latter in |sort l|, it must be the case that the indices are less than |pivot|.
 By |sound_cmp_expec_0|, it follows that the first |qs_U| term will produce no |(i, j)| comparisons, so the first |monoid_expec| term vanishes as well, leaving
@@
 To make proving the generated proof obligations possible, the types of |filter| and |partition| had to be sigma-decorated with modest length guarantees, because the opacity of the unspecified monad's |bind| operation prevented |Program| from being able to ``see'' the relation between the function's parameters and the arguments it passed in its recursive calls.
-\item The definitions produced by |Program| use clever recursion operators that translate a well-founded relation into a form that can be structurally recursed on.
-Consequently, corresponding induction principles are similarly clever.
-Using these induction principles for a function already defined with two layers of monadicity quickly became unwieldy.
-To deal with this, we defined a series of induction principles specialized for specific instances of |qs|, which put the recursive proof obligation in as simple terms as possible, with as little monadic indirection as possible, using among other things the observations regarding determinism of the comparison operation, as described in section \ref{finishing}.
+\item Using the induction principles corresponding to the definitions produced by |Program| for a function already defined with two layers of monadicity quickly became unwieldy.
+To deal with this, we defined a series of specialized induction principles for specific instances of |qs|, which put the recursive proof obligation in as simple terms as possible, with as little monadic indirection as possible, using among other things the observations regarding determinism of the comparison operation, as described in section \ref{finishing}.
 \end{itemize}

### 2008-08-29 paper/paper.tex
paper: Add list of expec identities.
darcs-hash:20080829024053-bab43-021b3ec5b841c04ed20f759d951344c517e88da9.gz

@@
 %format \/ = "\vee"
+%format <-> = "\leftrightarrow "
 %format MonoidMonadTrans.M = "\Varid{MMT}"
@@
 
+|expec| gives rise to a host of identities, including
+
+\begin{code}
+  0 <= expec f t
+  expec (fun x => f x + g x) t = expec f t + expec g t.
+  expec ((* c) fcmp f) = (* c) fcmp expec f
+
+  (forall x `In` t -> f x <= g x) -> expec f t <= expec g t.
+  (forall x `In` t -> f x = c) -> expec f t = c
+  (forall x, x `In` t -> f x = 0) <-> expec f t = 0
+
+  expec f (t >>= (ret fcmp g)) = expec (f fcmp g) t = expec f (ne_tree.map g t)
+\end{code}
+
 To form the monad with which we will instantiate |qs| for the main theorem, we now piggyback the additive monoid on |nat| onto |M_ne_tree| using |MMT|, and call the result |NDP| (for ``non-deterministically profiled''):
@@
 
-Like |expec|, |monoid_expec| gives rise to several identities. One identity that we would like to highlight is
+In addition to all the identities |monoid_expec| inherits from |expec|, it has some of its own.
+One identity worth highlighting states that if one transforms |M_ne_tree| using a monoid |m|, then for a monoid homomorphism |h| from |m| to the additive monoid on |nat|, |monoid_expec h| distributes over |bind|, provided that the expected monoid value of the right hand side does not depend on the computed value of the left hand side:
 \begin{code}
@@
     monoid_expec h (f >>= g) =
-      monoid_expec h f + monoid_expec h (g (snd (ne_tree.head f))),
+      monoid_expec h f + monoid_expec h (g (snd (ne_tree.head f))).
 \end{code}
-which states that if one transforms |M_ne_tree| using a monoid |m|, then for a monoid homomorphism |h| from |m| to the additive monoid on |nat|, |monoid_expec h| distributes over |bind|, provided that the expected monoid value of the right hand side does not depend on the computed value of the left hand side.
-Since |id| is a monoid homomorphism, |monoid_expec_plus| applies to |NDP| and |expec cost|.
-In section \ref{reduction}, we will use it with another monoid and homomorphism.
+Since |id| is a monoid homomorphism, |monoid_expec_plus| applies to |NDP| and |expec cost|. In section \ref{reduction}, we will use it with another monoid and homomorphism.
 
@@
 
-After rewriting once more, this time using\begin{code}expec_fcmp: forall f g t, expec (f fcmp g) t = expec f (ne_tree.map g t),\end{code}the goal becomes
+After rewriting with the last of the identities listed for |expec| in section \ref{nondetexpec}, the goal becomes
 \begin{code}
@@
 \end{code}}
-The last of the three subgoals thus generated states that there are no duplicate values in |IJ|, which trivially holds.
-The second subgoal generated is\begin{code}forall (i, j) `NotIn` IJ, expec (count (i, j)) (ne_tree.map fst (qs_U li)) = 0.\end{code} Rewriting this using |expec_fcmp| backwards, then rewriting the |expec| as a |monoid_expec|, and then generalizing the premise, results in the following property which is useful in its own right, as we will see later.
+It is trivially the case that |IJ| does not contain duplicate values.
+The second subgoal generated is\begin{code}forall (i, j) `NotIn` IJ, expec (count (i, j)) (ne_tree.map fst (qs_U li)) = 0.\end{code} Rewriting this using  the last of the identities listed for |expec| in section \ref{nondetexpec} in reverse, then rewriting the |expec| as a |monoid_expec|, and then generalizing the premise, results in the following property which is useful in its own right, as we will see later.
 \begin{code}sound_cmp_expec_0: forall i j li, (i `NotIn` li \/ j `NotIn` li) ->

### 2008-08-29 paper/paper.tex
optional arguments to \cite
darcs-hash:20080829130912-d98fc-6e36b8a61a0f1dc2eb78008f778c1ffea06c5cea.gz

@@
 
-Proofs of Quicksort's $O(n \log n)$ average-case complexity are included in many textbooks on computational complexity \cite{introtoalgos} \todo{more}.
+Proofs of Quicksort's $O(n \log n)$ average-case complexity are included in many textbooks on computational complexity \cite[for example, which we followed in this course of this research]{introtoalgos}.
+%%% \todo{more}.
 This paper documents what the authors believe to be the first fully formal machine-checked version of such a proof, developed using the Coq proof assistant \cite{coq}.

### 2008-08-29 paper/paper.tex
paper: separate abstract plus macro fiddling
abstract now comments out the bracket about translation between monoids;
comment macros now added, and a separate macros.sty file begun

darcs-hash:20080829140852-d98fc-d00f6165497425a064119bed20d3ce17fbbe15e7.gz

@@
 
+\usepackage{macros}
+
 \newcommand{\todo}[1]{\textcolor{magenta}{[TODO: #1]}}
@@
 \begin{abstract}
-As a case-study in machine-checked reasoning about the complexity of algorithms in type theory, we describe a proof of the average-case complexity of Quicksort in Coq.
-The proof attempts to follow a textbook development, at the heart of which lies a technical lemma about the behavior of the algorithm for which the original proof only gives an intuitive justification.
-Our formalization includes a fully formal treatment of this lemma, introducing a number of technical innovations:
-a shallow embedding of the algorithm as a monadically expressed functional program, instantiated with a variety of operation-counting monads that capture worst- and average case complexity of deterministic and non-deterministic programs, generalized to count in an arbitrary monoid (to support a key step in the proof where the expected comparison count is translated into the expected length of a recorded list of all comparisons);
-a small theory of expectation for non-deterministically computed values, featuring both general map-fusion like results, and specific counting arguments.
+  \input{abstract.ltx}
 \end{abstract}

### 2008-08-30 paper/paper.tex
paper: macro for author and title; bib: macro plus tidying
darcs-hash:20080830065529-d98fc-73bb4fcc53a3cc2f72f22b9a0748578496987489.gz

@@
 
+\newcommand{\theauthors}{James McKinna and Eelis van der Weegen}
+\newcommand{\thetitle}{A Machine-checked Proof of the Average-case Complexity of Quicksort in Coq}
+
 \usepackage{color}
-\usepackage[colorlinks,pdftitle={dus},pdfauthor={James McKinna and Eelis van der Weegen}]{hyperref}
+\usepackage[colorlinks,pdftitle={dus},pdfauthor=\theauthors]{hyperref}
 
@@
 
-\title{A Machine-checked Proof of the Average-case Complexity of Quicksort in Coq}
+\title{\thetitle}
 
@@
 
-\author{James McKinna and Eelis van der Weegen}
+\author{\theauthors}
 

### 2008-08-31 paper/paper.tex
paper/intro: Give a much more thorough overview of the structure of the paper and how it relates to the paper proof.
darcs-hash:20080831223900-bab43-71d643acbce50cf3c47296e2e3954d651e48bd42.gz

@@
 
-Proofs of Quicksort's $O(n \log n)$ average-case complexity are included in many textbooks on computational complexity \cite[for example, which we followed in this course of this research]{introtoalgos}.
-%%% \todo{more}.
+Proofs of the $O(n \log n)$ average-case complexity of Quicksort \cite{HoareQuick} are included in many textbooks on computational complexity \cite[for example]{introtoalgos}.
 This paper documents what the authors believe to be the first fully formal machine-checked version of such a proof, developed using the Coq proof assistant \cite{coq}.
 
-The formalization is based on the ``paper proof'' in \cite{introtoalgos}, but adds a great deal of theory simply to handle all the details that are omitted in that original.
-In particular, the latter takes key assertions about the algorithm's behavior for granted.
+The formalization is based on the ``paper proof'' in \cite{introtoalgos}, which consists of three parts.
+The first part argues that the total number of comparisons performed by the algorithm can be written as a sum of probabilities of comparisons between individual pairs of elements from the input list.
+The second part then derives from the algorithm a specific formula for those probabilities.
+Finally, the third and last part employs some analysis involving the harmonic series to derive the $O(n \log n)$ bound from the sum-of-probabilities.
+Thus, the first part essentially establishes an elementwise-comparison-probability lemma as an interface between the second and third part, proved by the former and used by the latter.
+
+Of these three parts, only the first two require reasoning about the actual behavior of the algorithm---the third part is just ordinary mathematics.
+While the original proof provides a thorough treatment of the third part, its treatment of the first two parts is informal in two major ways.
+
+First, it never actually justifies anything in terms of the algorithm's formal semantics.
 Indeed, it does not even formally define the algorithm in the first place.
+Instead, it relies on assertions which are taken to be intuitively true.
 While this practice is common and perfectly reasonable for paper proofs intended for human consumption, it is a luxury we do not afford ourselves.
-Hence, a large part of this paper is devoted to the representations we use to define the algorithms, and how they support complexity proofs.
-These representations take the form of a shallow monadic embedding, and depend critically on the type theoretic nature of Coq's calculus \cite{coq'art}.
 
-Section \ref{embed} introduces the basic idea of the embedding, and demonstrates its use by briefly proving Quicksort's quadratic worst-case complexity with it.
-In section \ref{nondetexpec}, the technique is expanded to handle the notion of the \emph{expected value} of non-deterministic programs, which then lets us state the main theorem in section \ref{statement}.
-Sections \ref{reduction} and \ref{finishing} detail the actual proof.
+Second, the original proof (implicitly) assumes that the input list does not contain any duplicate elements, which significantly simplifies its derivation of the formula for the probability of any two given elements being compared by the algorithm.
+We do not make this assumption.
+  % todo: bring this up later.
+
+The key to giving a proper formal treatment of both of these issues lies in the use of an appropriate representation of the algorithm, capable of capturing its computational behavior---specifically, its use of comparisons---in a way suitable for subsequent formal reasoning.
+The representation we use takes the form of a shallow monadic embedding, in which the algorithm is written as a monadically expressed functional program.
+This definition is then instantiated with refinements of operation-counting monads to make the number of comparisons performed observable.
+
+The embedding is introduced in section \ref{embed}, where we demonstrate its use by first defining a simple deterministic Quicksort implementation (which just selects the head of the input list as its pivot element, and uses two separate filter passes to partition the input list), and then proving that implementation's quadratic worst-case complexity.
+
+In preparation of the more complex average-case theorem, we then define (in section \ref{nondetexpec}) a nondeterministic Quicksort implementation which selects pivot elements at random, and which uses a single three-way partition pass instead of a two two-way filter passes.
+We also introduce a monad that combines operation counting with nondeterminism, supporting a formal definition of the notion of the \emph{expected} number of comparisons that are performed.
+
+Instantiating the nondeterministic implementation with this monad then lets us state the main theorem in section \ref{statement}.
+
+The next two sections detail the actual formalized proof.
+Section \ref{reduction} corresponds to the first part in the original proof, and shows how the main theorem can be split up into a lemma (stated in terms of another special-purpose monad) about the probability of comparisons being performed for pairs of input elements indexed a certain way, and a strictly numerical part.
+
+Since we were able to fairly directly transcribe the latter part from the paper proof, using the existing real-number theory in the Coq standard library with few complications and additions, we omit a discussion of it in this paper and refer the interested reader to the paper proof.
+
+Section \ref{finishing} shows the proof of the lemma mentioned above, and corresponds to the part in the original proof that derives the probability formula from the algorithm.
+Since this is the part where the original proof omits the most detail, and makes the assumption regarding duplicate elements, and where we really have to reason about the detailed behavior of the algorithm, it is by far the most involved part of the formalization.
+
 Section \ref{conclusion} ends with conclusions and final remarks.

### 2008-09-02 paper/paper.tex
paper: Many improvements to all sections except the introduction and the sections dealing with the proof.
darcs-hash:20080902214607-bab43-eedf50bb80530cc184d292794f423b124d507c2a.gz

@@
 % - mention that we don't do correctness proofs (although we can add them in the formalization, of course)
+% - decide whether to hide, ignore, or explain R/N differences. explaining costs precious space. ignoring is "begging the question". hiding is naughty.
+% - have a sentence somewhere hinting at the complications caused by duplicate elements (e.g., by pointing out simplifications that hold only with the assumption of no duplicate elements).
 
@@
 The formalization is based on the ``paper proof'' in \cite{introtoalgos}, which consists of three parts.
-The first part argues that the total number of comparisons performed by the algorithm can be written as a sum of probabilities of comparisons between individual pairs of elements from the input list.
-The second part then derives from the algorithm a specific formula for those probabilities.
-Finally, the third and last part employs some analysis involving the harmonic series to derive the $O(n \log n)$ bound from the sum-of-probabilities.
-Thus, the first part essentially establishes an elementwise-comparison-probability lemma as an interface between the second and third part, proved by the former and used by the latter.
+The first part argues that the total number of comparisons performed by the algorithm can be written as the sum of the expected number of comparisons between individual pairs of elements from the input list (ordered a certain way).
+The second part then derives from the algorithm a specific formula for this expectancy.
+Finally, the third and last part employs some analysis involving the harmonic series to derive the $O(n \log n)$ bound from the sum-of-expectancies.
+Thus, the first part essentially establishes a lemma about the expected number of comparisons between individual pairs of elements from the input list as an interface between the second and third part, proved by the former and used by the latter.
 
@@
 
-The embedding is introduced in section \ref{embed}, where we demonstrate its use by first defining a simple deterministic Quicksort implementation (which just selects the head of the input list as its pivot element, and uses two separate filter passes to partition the input list), and then proving that implementation's quadratic worst-case complexity.
+The embedding is introduced in section \ref{embed}, where we demonstrate its use by first defining a simple deterministic Quicksort implementation (which just selects the head of the input list as its pivot element, and uses two separate filter passes to partition the input list), and then defining an operation counting monad that lets us prove that implementation's quadratic worst-case complexity.
 
@@
 The next two sections detail the actual formalized proof.
-Section \ref{reduction} corresponds to the first part in the original proof, and shows how the main theorem can be split up into a lemma (stated in terms of another special-purpose monad) about the probability of comparisons being performed for pairs of input elements indexed a certain way, and a strictly numerical part.
+Section \ref{reduction} corresponds to the first part in the original proof, and shows how the main theorem can be split up into a lemma (stated in terms of another special-purpose monad) about the expected number of comparisons being performed for pairs of input elements indexed a certain way, and a strictly numerical part.
 
@@
 
-Section \ref{finishing} shows the proof of the lemma mentioned above, and corresponds to the part in the original proof that derives the probability formula from the algorithm.
-Since this is the part where the original proof omits the most detail, and makes the assumption regarding duplicate elements, and where we really have to reason about the detailed behavior of the algorithm, it is by far the most involved part of the formalization.
+Section \ref{finishing} shows the proof of the lemma mentioned above, and corresponds to the part in the original proof that derives the expectancy formula from the algorithm.
+Since this is the part where the original proof omits the most detail, and makes the assumption regarding duplicate elements, and where we really have to reason in detail about the behavior of the algorithm, it is by far the most involved part of the formalization.
 
@@
 
-Before the main theorem can even be formally stated, let alone proved, a formal definition of the algorithm is needed, expressed in a programming language with well-defined semantics, suitable for subsequent formal reasoning.
-Here, we exploit the computational nature of the type theory on which Coq is based.
-Rather than defining the syntax and semantics of a programming language from scratch (this would be a ``deep'' embedding) like we would have to do in, say, ZFC mathematics, we instead use the built-in lambda calculus as a functional programming language, and write the algorithm as a specially adorned lambda term.
-More specifically, we express the algorithm monadically, with both the monad and the counted operation (in our case, the elementwise comparison---the usual complexity metric for sorting algorithms) made into parameters.
+% todo: less pompous here \/
+
+As stated before, the key to giving a proper formal treatment of those parts of the proof for which the original contents itself with appeals to intuition, lies in the use of an appropriate representation of the algorithm.
+Indeed, we cannot even formally state the main theorem until we have both an algorithm definition and the means to denote its use of comparisons (the usual complexity metric for sorting algorithms).
+
+Since we are working in Coq, we already have at our disposal a full functional programming language, in the form of Coq's pCIC \cite{coq'art}.
+However, just writing the algorithm as an ordinary Coq function would not let us observe its use of comparisons.
+We make those uses observable by writing the algorithm as a monadically expressed function, parameterizing it on both the monad itself and on the comparison operation.
 Expressed this way, a deterministic quicksort that simply selects the head of the input list as its pivot element, and uses two simple filter passes to partition the input list, looks as follows:
@@
 For a general introduction to monadic programming and monad laws, see \cite{wadler93monads}.
-The notation |x <- y; z|, shorthand for |y >>= \x -> z| (where |x| may occur in |z|) is known as ``do-notation''.
+The notation ``|x <- y; z|'', shorthand for |y >>= \x => z| (where |x| may occur in |z|) is known as ``do-notation''.
 
@@
 
-For reasons that will become clear in later sections, we compose the monad we will instantiate the above definitions with using a monad transformer called |MMT| (for ``monoid monad transformer''), which piggybacks a monoid onto an existing monad.
+For reasons that will become clear in later sections, we compose the monad we will instantiate the above definitions with using a monad transformer called |MMT| (for ``monoid monad transformer''), which piggybacks a monoid onto an existing monad by pairing.
 
@@
 
-We use |MMT| to piggyback the additive monoid on |nat| onto the identity monad, and lift elementwise comparison into the resulting monad, which we call |SP| (for ``simply-profiled'').
+We now use |MMT| to piggyback the additive monoid on |nat| onto the identity monad, and lift elementwise comparison into the resulting monad, which we call |SP| (for ``simply-profiled'').
 \begin{code}
@@
 
-\begin{propo}|qs_worst: forall l, cost (qs_SP l) <= sqrd (length l).|\end{propo}
-  \todo{get rid of period after ``Proposition'' above}
+\begin{propo}|qs_worst: forall l, cost (qs_SP l) <= sqrd (length l).|\footnote{(We don't use big-$O$ notation for this simple statement, as it would only obfuscate.
+Big-$O$ complexity is discussed in section \ref{statement}.)}
+\end{propo}
+
 \begin{proof}
@@
 The induction hypothesis applies to the recursive |qs_SP| calls.
-Furthermore,
+Furthermore, by induction on |t|, we can easily prove
   \begin{code}
   length (result (filter (le pivot) t)) +
-  length (result (filter (gt pivot) t)) <= length t
+  length (result (filter (gt pivot) t)) <= length t,
   \end{code}
-  can easily be proved by induction on |t|, because the two predicates filtered on are mutually exclusive. Abstracting |filter (le pivot) t| and its |gt| counterpart, this leaves
+  because the two predicates filtered on are mutually exclusive. Abstracting |filter (le pivot) t| and its |gt| counterpart, this leaves
   \begin{code}
@@
 
-We now expand the machinery in preparation of the average-case complexity proof.
+We now expand the technique in preparation of the average-case complexity proof.
 
@@
 
-The Quicksort algorithm used by the average-case proof we formalize differs from the one presented in the last section in two ways: it is nondeterministic, and uses a single three-way partition pass instead of two two-way filter passes.
-Combined, these two traits ensure that the $O(n \log n)$ average-case property holds not just averaged over all input lists, but for each individual input list as well; nondeterministic pivot selection avoids the pathological cases any deterministic pivot selection strategy inevitably suffers, while a single three-way partition pass avoids the pathological quadratic case that two two-way filter passes suffer for an input list consisting of elements all of the same value.
-This greatly simplifies things, because it means that the global bound will follow immediately if it is proved for an arbitrary input.
+The Quicksort algorithm used by the average-case proof we formalize differs from the one presented in the last section in two ways.
+First, it selects a pivot element nondeterministically, thereby avoiding the pathological quadratic cases any deterministic pivot selection strategy inevitably suffers.
+Second, it uses a single three-way partition pass instead of two two-way filter passes, to avoid the pathological quadratic case that two two-way filter passes suffer for an input list consisting of elements all of the same value.
 
-Hence, we define
+Combined, these two traits ensure that the $O(n \log n)$ average-case bound holds not just averaged over all input lists, but for each individual input list as well.
+This greatly simplifies things, because it means that the global bound can be proved simply by proving it as a bound of the expected number of comparisons performed for an arbitrary input.
+
+Hence, we now modify our |qs| definition to use a three-way partition pass, and to use nondeterministic pivot selection.
+Following that, we will compose a new monad (again using |MMT|, but this time transforming a nondeterminism monad) that the new definition can be instantiated with to capture its \emph{expected} number of comparisons.
+
+First, we define |partition|.
+Instead of a two-way comparison operation producing a monadic |bool|, it takes a three-way comparison operation producing a monadic |comparison|, which is an enumeration with values |Lt|, |Eq|, and |Gt|.
+|partition| constructs a |Partitioning T|, which is simply a function of type |comparison -> list T|.
+We might equally well have used an inductive definition of |Partitioning T| as a record containing three lists.
+% at least until i remember why i used a function.
 
@@
     end.
+\end{code}}
 
+Next, we redefine |qs| to use |partition|, and have it take as an additional parameter a |pick| operation, representing nondeterministic selection of an element of a non-empty list of choices.
+(An |ne_list T| is a non-empty list of |T|'s.)
+
+\begin{code}
   Variable pick: forall (A: Set), ne_list A -> M A.
@@
     end.
-\end{code}}
-Here, |comparison| is an enumeration with values |Lt|, |Eq|, and |Gt|.
-A |Partitioning T| is a function of type |comparison -> list T|.
-An |ne_list T| is a non-empty list of |T|'s.
+\end{code}
 The functions |nth| and |remove| select and remove the $n$th element of a list, respectively.
@@
 \end{code}
-(We use non-empty trees because we do not consider computations that produce no values at all, and using potentially empty trees would complicate the definition of a tree's average value below.)
+We use non-empty trees because we do not consider partial functions, and using potentially empty trees would complicate the definition of a tree's average value below.
+This is also why we used |ne_list| for |pick|.
 
@@
 
-Because we are often interested in the expected value of a measure of the output of a program that does not return numbers, we further define
+To denote the expected value of a \emph{measure} of the output of a program, we define
 
@@
 
-To form the monad with which we will instantiate |qs| for the main theorem, we now piggyback the additive monoid on |nat| onto |M_ne_tree| using |MMT|, and call the result |NDP| (for ``non-deterministically profiled''):
+To form the monad with which we will instantiate |qs| for the main theorem, we now piggyback the additive monoid on |nat| onto |M_ne_tree| using |MMT|, and call the result |NDP| (for ``nondeterministically profiled''):
 \begin{code}
@@
 \end{code}
+We can now denote the expected number of comparisons of a |qs_NDP| application by |expec cost (qs_NDP l)|, and will use this in our statement of the main theorem in the next section.
 
-When working with computations in a monad formed by transforming |M_ne_tree| using |MMT|, we are often interested in the expected value of a function of the piggybacked monoid.
-For this, we introduce |monoid_expec|.
+But before we do so, we define a slight refinement of |expec| that specifically observes the monoid component of computations in monads formed by transforming |M_ne_tree| using |MMT| (like |NDP|).
 \begin{code}
@@
 In addition to all the identities |monoid_expec| inherits from |expec|, it has some of its own.
-One identity worth highlighting states that if one transforms |M_ne_tree| using a monoid |m|, then for a monoid homomorphism |h| from |m| to the additive monoid on |nat|, |monoid_expec h| distributes over |bind|, provided that the expected monoid value of the right hand side does not depend on the computed value of the left hand side:
-\begin{code}
+One identity states that if one transforms |M_ne_tree| using a monoid |m|, then for a monoid homomorphism |h| from |m| to the additive monoid on |nat|, |monoid_expec h| distributes over |bind|, provided that the expected monoid value of the right hand side does not depend on the computed value of the left hand side:
+
+\parbox{\textwidth}{\begin{code}
   monoid_expec_plus: forall (m: Monoid) (h: m -> nat) (A B: Set)
@@
       monoid_expec h f + monoid_expec h (g (snd (ne_tree.head f))).
-\end{code}
-Since |id| is a monoid homomorphism, |monoid_expec_plus| applies to |NDP| and |expec cost|. In section \ref{reduction}, we will use it with another monoid and homomorphism.
+\end{code}}
+
+Since |id| is a monoid homomorphism, |monoid_expec_plus| applies to |NDP| and |expec cost|. In section \ref{reduction}, we will use |monoid_expec_plus| with another monoid and homomorphism.
 
@@
 
-Now that we have a solid definition of nondeterministic, instrumented quicksort, as well as the means to express the expected values of nondeterministic programs, the last thing needed before the main theorem can be stated, is the notion of big-O complexity.
+The last thing needed before the main theorem can be stated, is the notion of big-O complexity.
 The definition we use is a standard textbook definition, except that we work with a separate measure function parameter:
@@
 \end{code}
-Thanks to the property discussed at the start of the previous section, the above follows as a corollary from the (slightly stronger) statement that
+Thanks to the property discussed at the start of the previous section, % perhaps too far back
+the above follows as a corollary from the (slightly stronger) statement that
 \begin{code}
@@
 There, we use vectors (lists whose size is part of their type) and bounded natural numbers in many places instead.
-Using these substantially reduces the amount of |i < length l| proofs that need to be produced, converted, and passed around, but the issue is still far from painless.
+Using these substantially reduces the amount of |i < length l| proofs that need to be produced, converted, and passed around, but this solution is still far from painless.
 \item Using the |Program| facility to deal with quicksort's non-structural recursion was not completely as trivial as we made it out to be.
@@
 
-In conclusion, we have found the shallow monadic embedding to be a straightforward, clean, and solid basis for reasoning about an algorithm's computational complexity.
-Calling it an embedding almost seems misplaced, since the algorithm definitions are essentially just monadically expressed functional programs in Coq's calculus.
+The shallow monadic embedding provides a simple but effective representation of the algorithm.
+The parameterization of the monad makes the representation very flexible, allowing a single definition to be instantiated either with basic monads (like the identity monad or bare nondeterminism monads) to get a non-instrumented version suitable for extraction and correctness proofs, or with monads that use |MMT| to add operation-counting, to support our complexity proofs.
+
+Furthermore, since the shallowness of the embedding lets us re-use all standard Coq data types and facilities, including the powerful |Program Fixpoint| command, the actual algorithm definition itself is reasonably clean.
+
+The |MMT| monoid monad transformer provided the ``glue'' to add instrumentation onto a base monad.
+Its use of a monoid to represent the operation ``counter'' let us use a single set of lemmas (about |monoid_expec|) for |NDP| and |U|, as both were instances of the |M_ne_tree| monad transformed with |MMT|, differing only in the monoid used.
+Indeed, the relative ease with which we could transform the expectancy lemma into the |U| monad, a monad highly specialized for the specific order-index based counting scheme we used, further attests to the flexibility of the combination of a monadic representation and the |MMT| monad transformer.
+
+\todo{Paragraph saying that the formalized proof is faithful to the original proof, except for the part where we dropped the duplicate elements assumption.
+The purely mathematical part could be directly transcribed.
+The only theory we had to add served to support a formal treatment of the parts that the original clearly took for granted.}
+
+% old passage that may yet be useful in some form or another:
+%    Calling it an embedding almost seems misplaced, since the algorithm definitions are essentially just monadically expressed functional programs in Coq's calculus.
 Furthermore, since monads and monad transformers have been used in functional programming for decades, the basic idea should be immediately clear to any functional programmer.
@@
 
-Quicksort turned out to be a good stress test for the technique, showing that it copes well with nondeterminism and nontrivial average-case bounds.
-The |U| monad shows that custom profiling monads can be defined to analyse very algorithm-specific traits for a specific part of a larger complexity proof, which was crucial for this particular proof.
-
-For future work, a logical next challenge is to try and formalize a (likely more complex) proof that the $O(n \log n)$ bound holds also for deterministic quicksorts (with sensible pivot selection strategies).
+For future work, an interesting next challenge is to try and formalize a (likely more complex) proof that the $O(n \log n)$ bound holds also for deterministic quicksorts (with sensible pivot selection strategies).
 

### 2008-09-21 paper/paper.tex
paper: Fix lots of things (layout, code formatting, code itself, wording). Truncate things. Change conclusion somewhat.  src: Simplify some things. Fix many small todo's. Add some stuff for reasoning about the simpler quicksort definitions, used by the correctness and equivalence proofs (to be added soon).
darcs-hash:20080921225058-bab43-f34d7b096a00b3d10121dae840ec9d423155ba05.gz

@@
 \usepackage{color}
-\usepackage[colorlinks,pdftitle={dus},pdfauthor=\theauthors]{hyperref}
+\usepackage[colorlinks,pdftitle=\thetitle,pdfauthor=\theauthors]{hyperref}
 
@@
 
+%format sp = "\ "
+
+%format == = "= "
+%format /= = "\ne "
+  % Without these, lhs2tex uses three-line equal signs.
+
 %format fcmp = "\circ"
@@
 %format NatAddMonoid = (nat, 0, +)
+
+%format { = "\hspace{0.9mm}\{"
+  % so that parameters declared with {...} are spaced like those declared with (...)
 
@@
 %format M_NDP
-%format le_NDP
+%format cmp_NDP
 %format qs_NDP
@@
 % - check that names are spelled and encoded right in bibliography
-% - lhs2tex shows parameters declared with {...} differently from those declared with (...)
-% - mention that we don't do correctness proofs (although we can add them in the formalization, of course)
 % - decide whether to hide, ignore, or explain R/N differences. explaining costs precious space. ignoring is "begging the question". hiding is naughty.
@@
 
-\newcommand{\todo}[1]{\textcolor{magenta}{[TODO: #1]}}
+\arrayhs % i.e. no page breaks in code blocks
 
@@
 \section{Introduction}
+\label{intro}
 
@@
 The formalization is based on the ``paper proof'' in \cite{introtoalgos}, which consists of three parts.
-The first part argues that the total number of comparisons performed by the algorithm can be written as the sum of the expected number of comparisons between individual pairs of elements from the input list (ordered a certain way).
-The second part then derives from the algorithm a specific formula for this expectancy.
-Finally, the third and last part employs some analysis involving the harmonic series to derive the $O(n \log n)$ bound from the sum-of-expectancies.
-Thus, the first part essentially establishes a lemma about the expected number of comparisons between individual pairs of elements from the input list as an interface between the second and third part, proved by the former and used by the latter.
+The first part shows that the total number of comparisons performed by the algorithm (the usual complexity metric for sorting algorithms) can be written as a sum of expected comparison counts for individual pairs of input list elements.
+The second part derives from the algorithm a specific formula for this expectancy.
+The third and last part employs some analysis involving the harmonic series to derive the $O(n \log n)$ bound from the sum-of-expectancies.
 
-Of these three parts, only the first two require reasoning about the actual behavior of the algorithm---the third part is just ordinary mathematics.
+Of these three parts, only the first two involve the actual algorithm itself---the third part is strictly numerical.
 While the original proof provides a thorough treatment of the third part, its treatment of the first two parts is informal in two major ways.
@@
 First, it never actually justifies anything in terms of the algorithm's formal semantics.
-Indeed, it does not even formally define the algorithm in the first place.
-Instead, it relies on assertions which are taken to be intuitively true.
+Indeed, it does not even formally define the algorithm in the first place, relying instead on assertions which are taken to be intuitively true.
 While this practice is common and perfectly reasonable for paper proofs intended for human consumption, it is a luxury we do not afford ourselves.
 
-Second, the original proof (implicitly) assumes that the input list does not contain any duplicate elements, which significantly simplifies its derivation of the formula for the probability of any two given elements being compared by the algorithm.
+Second, the original proof (implicitly) assumes that the input list does not contain any duplicate elements, which significantly simplifies its derivation of the formula for the expected comparison count for pairs of individual input list elements.
 We do not make this assumption.
-  % todo: bring this up later.
+  % todo: perhaps bring this up later.
 
-The key to giving a proper formal treatment of both of these issues lies in the use of an appropriate representation of the algorithm, capable of capturing its computational behavior---specifically, its use of comparisons---in a way suitable for subsequent formal reasoning.
+The key to giving a proper formal treatment of both of these aspects lies in the use of an appropriate representation of the algorithm, capable of capturing its computational behavior---specifically, its use of comparisons---in a way suitable for subsequent formal reasoning.
 The representation we use takes the form of a shallow monadic embedding, in which the algorithm is written as a monadically expressed functional program.
-This definition is then instantiated with refinements of operation-counting monads to make the number of comparisons performed observable.
+This definition is then instantiated with refinements of operation-counting monads to make the comparison count observable.
 
-The embedding is introduced in section \ref{embed}, where we demonstrate its use by first defining a simple deterministic Quicksort implementation (which just selects the head of the input list as its pivot element, and uses two separate filter passes to partition the input list), and then defining an operation counting monad that lets us prove that implementation's quadratic worst-case complexity.
+The embedding is introduced in section \ref{embed}, where we demonstrate its use by first giving a simple deterministic monadic Quicksort definition, and then instantiating it with a simple operation counting monad that lets us prove its quadratic worst-case complexity.
 
-In preparation of the more complex average-case theorem, we then define (in section \ref{nondetexpec}) a nondeterministic Quicksort implementation which selects pivot elements at random, and which uses a single three-way partition pass instead of a two two-way filter passes.
-We also introduce a monad that combines operation counting with nondeterminism, supporting a formal definition of the notion of the \emph{expected} number of comparisons that are performed.
-
-Instantiating the nondeterministic implementation with this monad then lets us state the main theorem in section \ref{statement}.
+For the purposes of the more complex average-case theorem, we then give (in section \ref{nondetexpec}) a potentially-nondeterministic monadic Quicksort definition, and compose a monad that combines operation counting with nondeterminism, supporting a formal definition of the notion of the \emph{expected} comparison count, with which we state the main theorem in section \ref{statement}.
 
 The next two sections detail the actual formalized proof.
-Section \ref{reduction} corresponds to the first part in the original proof, and shows how the main theorem can be split up into a lemma (stated in terms of another special-purpose monad) about the expected number of comparisons being performed for pairs of input elements indexed a certain way, and a strictly numerical part.
+Section \ref{reduction} corresponds to the first part in the original proof described above, and shows how the main theorem can be split up into a lemma (stated in terms of another specialized monad) giving a formula for the expected comparison count for individual pairs of input element, and a strictly numerical part.
+Since we were able to fairly directly transcribe the latter part from the paper proof, using the existing real number theory in the Coq standard library with few complications and additions, we omit a discussion of it in this paper and refer the interested reader to the paper proof.
 
-Since we were able to fairly directly transcribe the latter part from the paper proof, using the existing real-number theory in the Coq standard library with few complications and additions, we omit a discussion of it in this paper and refer the interested reader to the paper proof.
-
-Section \ref{finishing} shows the proof of the lemma mentioned above, and corresponds to the part in the original proof that derives the expectancy formula from the algorithm.
+Section \ref{finishing} finishes the proof by proving the lemma about the expected comparison count for individual input list elements.
 Since this is the part where the original proof omits the most detail, and makes the assumption regarding duplicate elements, and where we really have to reason in detail about the behavior of the algorithm, it is by far the most involved part of the formalization.
@@
 
-% todo: less pompous here \/
-
 As stated before, the key to giving a proper formal treatment of those parts of the proof for which the original contents itself with appeals to intuition, lies in the use of an appropriate representation of the algorithm.
-Indeed, we cannot even formally state the main theorem until we have both an algorithm definition and the means to denote its use of comparisons (the usual complexity metric for sorting algorithms).
+Indeed, we cannot even formally state the main theorem until we have both an algorithm definition and the means to denote its use of comparisons.
 
@@
 
-  Fixpoint filter (c: T -> M bool) (l: list T): M (list T) :=
-    match l with
-    | nil => ret nil
-    | h :: t =>
-      t' <- filter c t;
-      b <- c h;
-      ret (if b then h :: t' else t')
-    end.
-
   Program Fixpoint qs (l: list T) {measure length l}: M (list T) :=
@@
     | pivot :: t =>
-        lower <- filter (gt pivot) t >>= qs;
-        upper <- filter (le pivot) t >>= qs;
+        lower <- filterM (gt pivot) t >>= qs;
+        upper <- filterM (le pivot) t >>= qs;
         ret (lower ++ pivot :: upper)
@@
 By instantiating the above definitions with the right monad, we can transparently insert comparison-counting instrumentation into the algorithm, which will prove to be sufficient to let us reason about its complexity.
-But before we do so, let us note that if the above definitions are instead instantiated with the identity monad and an ordinary elementwise comparison on |T|, then the monadic scaffolding melts away, and the result is equivalent to an ordinary non-instrumented, non-monadic version, suitable for extraction and correctness proofs.
+But before we do so, let us note that if the above definitions are instead instantiated with the identity monad and an ordinary elementwise comparison on |T|, then the monadic scaffolding melts away, and the result is equivalent to an ordinary non-instrumented, non-monadic version, suitable for extraction and correctness proofs (which are included in the formalization for completeness).
 This means that while we will instantiate the definitions with less trivial monads to support our complexity proofs, we can take some comfort in knowing that the object of those proofs is, in a very concrete sense, the actual Quicksort algorithm (as one would write it in a functional programming language), rather than some idealized model thereof.
 
-For reasons that will become clear in later sections, we compose the monad we will instantiate the above definitions with using a monad transformer called |MMT| (for ``monoid monad transformer''), which piggybacks a monoid onto an existing monad by pairing.
+For reasons that will become clear in later sections, we compose the monad we will instantiate the above definitions with using a monad transformer |MMT| (for ``monoid monad transformer''), which piggybacks a monoid onto an existing monad by pairing.
 
@@
 \end{code}
-Defining |cost| and |result| as the first and second projection, respectively, we have the identities
-\begin{code}
-  forall x, cost (ret_SP x) = 0,
-  forall x f, cost (bind_SP x f) = cost x + cost (f (result x),
-  forall x y, cost (le_SP x y) = 1.
-\end{code}
+
+%format `infixbind_SP` = >>="_{SP}"
+
+Defining |cost| and |result| as the first and second projection, respectively, we trivially have identities such as |cost (ret_SP x) = 0|, |cost (x `infixbind_SP` f) = cost x + cost (f (result x))|, and |cost (le_SP x y) = 1|.
 This very modest amount of machinery is sufficient for a straightforward proof of Quicksort's quadratic worst-case complexity.
 
-\begin{propo}|qs_worst: forall l, cost (qs_SP l) <= sqrd (length l).|\footnote{(We don't use big-$O$ notation for this simple statement, as it would only obfuscate.
-Big-$O$ complexity is discussed in section \ref{statement}.)}
+\begin{propo}|qs_worst: forall l, cost (qs_SP l) <= sqrd (length l).|\footnote{(We do not use big-O notation for this simple statement, as it would only obfuscate.
+Big-O complexity is discussed in section \ref{statement}.)}
 \end{propo}
@@
 For an empty input list, we have |cost (qs_SP nil) = cost (ret nil) = 0 <= sqrd (length l)|.
-For a non-empty input list |(pivot :: t)|, the cost decomposes into
+For a non-empty input list |h :: t|, the cost decomposes into
   \begin{code}
-    cost (filter (le pivot) t) + cost (qs_SP (result (filter (lt pivot) t))) +
-    cost (filter (gt pivot) t) + cost (qs_SP (result (filter (gt pivot) t))) +
-    cost (ret (result (qs_SP (result (filter (lt pivot) t))) ++
-     pivot :: result (qs_SP (result (filter (gt pivot) t))))).
+    cost (filter (le h) t) + cost (qs_SP (result (filter (lt h) t))) +
+    cost (filter (gt h) t) + cost (qs_SP (result (filter (gt h) t))) +
+    cost (ret (result (qs_SP (result (filter (lt h) t))) ++
+     h :: result (qs_SP (result (filter (gt h) t))))).
   \end{code}
@@
   \begin{code}
-  length (result (filter (le pivot) t)) +
-  length (result (filter (gt pivot) t)) <= length t,
+  length (result (filter (le h) t)) + length (result (filter (gt h) t)) <= length t,
   \end{code}
-  because the two predicates filtered on are mutually exclusive. Abstracting |filter (le pivot) t| and its |gt| counterpart, this leaves
+  because the two predicates filtered on are mutually exclusive. Abstracting the filter terms as |flt| and |flt'|, this leaves
   \begin{code}
-    forall (t flt flt': list T), length flt + length flt' <= length t ->
+    length flt + length flt' <= length t ->
       length t + sqrd (length flt) + length t + sqrd (length flt') + 0 <= sqrd (S (length t)),
@@
 First, it selects a pivot element nondeterministically, thereby avoiding the pathological quadratic cases any deterministic pivot selection strategy inevitably suffers.
-Second, it uses a single three-way partition pass instead of two two-way filter passes, to avoid the pathological quadratic case that two two-way filter passes suffer for an input list consisting of elements all of the same value.
+Second, it uses a single three-way partition pass instead of two two-way filter passes, thereby avoiding the pathological quadratic case that two two-way filter passes suffer for an input list whose elements are all of the same value.
 
-Combined, these two traits ensure that the $O(n \log n)$ average-case bound holds not just averaged over all input lists, but for each individual input list as well.
-This greatly simplifies things, because it means that the global bound can be proved simply by proving it as a bound of the expected number of comparisons performed for an arbitrary input.
+Combined, these two traits ensure that the $O(n \log n)$ average-case bound holds not just averaged over all possible input lists, but for each individual input list as well.
+This greatly simplifies things, because it means that if we prove the bound for an arbitrary input, the global bound immediately follows.
 
 Hence, we now modify our |qs| definition to use a three-way partition pass, and to use nondeterministic pivot selection.
-Following that, we will compose a new monad (again using |MMT|, but this time transforming a nondeterminism monad) that the new definition can be instantiated with to capture its \emph{expected} number of comparisons.
+Following that, we will compose a new monad (again using |MMT|, but this time transforming a nondeterminism monad) that the new definition can be instantiated with to capture its \emph{expected} comparison count.
 
@@
 Instead of a two-way comparison operation producing a monadic |bool|, it takes a three-way comparison operation producing a monadic |comparison|, which is an enumeration with values |Lt|, |Eq|, and |Gt|.
-|partition| constructs a |Partitioning T|, which is simply a function of type |comparison -> list T|.
-We might equally well have used an inductive definition of |Partitioning T| as a record containing three lists.
-% at least until i remember why i used a function.
 
-\parbox{\textwidth}{\begin{code}
+\begin{code}
   Variables (T: Set) (M: Monad) (cmp: T -> T -> M comparison).
 
-  Fixpoint partition (pivot: T) (l: list T): M (Partitioning T) :=
+  Fixpoint partition (t: T) (l: list T): M (comparison -> list T) :=
     match l with
-    | nil => ret emptyPartitioning
-    | h :: t =>
-        b <- cmp h pivot;
-        tt <- partition pivot t ;
-        ret (addToPartitioning b h tt)
+    | nil => ret (const nil)
+    | h :: l' =>
+        c <- cmp h t; f <- partition t l';
+        ret (fun c' => if c == c' then h :: f c' else f c')
     end.
-\end{code}}
+\end{code}
 
@@
 \begin{code}
-  Variable pick: forall (A: Set), ne_list A -> M A.
+  Variable pick: forall A: Set, ne_list A -> M A.
 
@@
 
-Nondeterminism can now be achieved by instantiating these definitions with a suitable monad, along with a corresponding |pick| operation.
-A deterministic, non-instrumented version can still be obtained, simply by instantiating the definitions with the identity monad and a corresponding deterministic |pick| operation, like |head|.
+Nondeterminism can now be emulated by instantiating these definitions with a suitable monad and |pick| operation.
+A deterministic, non-instrumented version can still be obtained, simply by using the identity monad and a deterministic |pick| operation, like |head|.
 
-Let us now consider what kind of nondeterminism monad would be suitable for reasoning about the expected value of a nondeterministic program.
-The list monad is commonly used to emulate nondeterministic computation.
-With the list monad, the program
+Let us now consider what kind of nondeterminism monad would be suitable for reasoning about the expected value of a nondeterministic program like
 \begin{code}
-  x <- pick [0, 1]; if x = 0 then ret 0 else pick [1, 2],
+  x <- pick [0, 1]; if x = 0 then ret 0 else pick [1, 2].
 \end{code}
+When executed in the list monad (commonly used to emulate nondeterministic computation), this program
 produces |[0, 1, 2]| as its list of possible outcomes.
@@
     end.
-
+\end{code}
+\begin{code}
   Definition M_ne_tree: Monad := Build_Monad ne_tree bind_ne_tree ret_ne_tree.
@@
 
-|expec| gives rise to a host of identities, including
+|expec| gives rise to a host of identities, such as
 
@@
 
-  Definition le_NDP (x y: T): M bool := ret_ne_tree (1, le x y).
+  Definition cmp_NDP (x y: T): M bool := ret_ne_tree (1, cmp x y).
 
-  Definition qs_NDP := qs M_NDP le_NDP (lift pick_ne_tree).
+  Definition qs_NDP := qs M_NDP cmp_NDP (lift pick_ne_tree).
 \end{code}
-We can now denote the expected number of comparisons of a |qs_NDP| application by |expec cost (qs_NDP l)|, and will use this in our statement of the main theorem in the next section.
+We can now denote the expected comparison count for a |qs_NDP| application by |expec cost (qs_NDP l)|, and will use this in our statement of the main theorem in the next section.
 
@@
 \end{code}
-Since |cost = fst|, we have |forall t, expec cost t = monoid_expec id t|.
+Since |cost = fst|, we have |expec cost t = monoid_expec id t|.
 
@@
 
-\parbox{\textwidth}{\begin{code}
-  monoid_expec_plus: forall (m: Monoid) (h: m -> nat) (A B: Set)
+\begin{code}
+  monoid_expec_plus: forall (m: Monoid) (h: m -> NatAddMonoid),
+    monoid_homo h -> forall (A B: Set)
     (f: MMT m NeTreeMonad A) (g: A -> MMT m NeTreeMonad B):
@@
       monoid_expec h f + monoid_expec h (g (snd (ne_tree.head f))).
-\end{code}}
+\end{code}
 
@@
 
-The last thing needed before the main theorem can be stated, is the notion of big-O complexity.
-The definition we use is a standard textbook definition, except that we work with a separate measure function parameter:
+The last thing needed before the main theorem can be stated, is the notion of big-O complexity. We use a standard textbook definition, except that we incorporate a measure function parameter:
 
-\parbox{\textwidth}{\begin{code}
-  Definition measured_bigO (X: Set)
-    (m: X -> nat) (f: X -> R) (g: nat -> R): Prop
-      := exists c, exists n, forall x, n <= m x -> f x <= c * g (m x).
+\begin{code}
+  Definition bigO (X: Set) (m: X -> nat) (f: X -> R) (g: nat -> R): Prop
+      := exists c n, forall x, n <= m x -> f x <= c * g (m x).
 
-  Notation {-"``"-}over m ,{-"\ "-} f =O( g ){-"\text{''''}"-} := (measured_bigO m f g).
-\end{code}}
+  Notation {-"``"-}over m,{-"\ "-} f =O( g ){-"\text{''''}"-} := bigO m f g.
+\end{code}
 We now state the main theorem.
 \begin{code}
-  Theorem qs_avg: over length,
-    expec cost fcmp qs_NDP =O(fun n => INR (n * log2ceil n)).
+  Theorem qs_avg: over length, expec cost fcmp qs_NDP =O(fun n => INR (n * log2ceil n)).
 \end{code}
 Thanks to the property discussed at the start of the previous section, % perhaps too far back
-the above follows as a corollary from the (slightly stronger) statement that
+|qs_avg| follows as a corollary from the stronger statement that
 \begin{code}
-  Theorem qs_expec_cost:
   forall l, expec cost (qs_NDP l) <= 2 * length l * S (log2ceil (length l)),
 \end{code}
-the proof of which is described in the remainder of this paper.
+the proof of which is described in the next two sections.
 
@@
 
-The main idea in the proof is to reduce |qs_expec_cost| to a statement about the expected number of comparisons between specific pairs of elements from the input list.
-
-If $X \equiv X_{I_0} \ldots X_{I_{n-1}}$ is the input list, with $I$ a permutation of |[0 ...
-n-1]| such that $X_0 \ldots X_n$ is sorted, then the expected number of comparisons between any $X_i$ and $X_j$ with $i < j$ is at most $2 / S (j - i)$.
-In other words, the expected number of comparisons between two list elements is bounded by a simple function of the number of list elements that separate the two in the sort order.
+As described in the introduction, the key ingredient in the proof is a lemma giving a formula for the expected comparison count for individual pairs of input list elements, indexed a certain way.
+More specifically, if $X \equiv X_{I_0} \ldots X_{I_{n-1}}$ is the input list, with $I$ a permutation of |[0 ...
+n-1]| such that $X_0 \ldots X_n$ is sorted, then the expected comparison count for any $X_i$ and $X_j$ with $i < j$ is at most $2 / S (j - i)$.
+In other words, the expected comparison count for two input list elements is bounded by a simple function of the number of list elements that separate the two in the sort order.
 We prove this fact in the next section, but first show how |qs_expec_cost| follows from it.
 
-Combined with the observation that the total expected number of comparisons equals the sum of the expected numbers of comparisons for each |(i, j)| in $\Varid{IJ} := \{ (i, j) \in [0, |length l|) \mid i < j \}$, the property described above suggests breaking up the inequality into
+Combined with the observation that the total expected comparison count ought to equal the sum of the expected comparison count for each individual pair of input elements, the property described above suggests breaking up the inequality into
 \begin{code}
-  expec cost (qs_NDP l) <= {-"\displaystyle{\sum_{(i, j) \in \Varid{IJ}}\frac{2}{S (j - i)}}"-} <= 2 * length l * S (log2ceil (length l)).
+  expec cost (qs_NDP l) <= {-"\displaystyle{\sum_{(i, j) \in \Varid{IJ}}\frac{2}{S (j - i)}}"-} <= 2 * length l * S (log2ceil (length l)),
 \end{code}
-Proving the right inequality requires a bit of analysis involving the harmonic series.
-This part of the proof could be fairly directly transcribed from the paper proof, using the existing real-number theory in the Coq standard library, with few complications and additions.
-We refer the interested reader to the paper proof.
+where $\Varid{IJ} := \{ (i, j) \in [0, |length l|) \mid i < j \}$.
 
-The left inequality is the challenging one. To bring it closer to the index summation, we first rewrite
-\begin{code}
-  expec cost (qs_NDP l) = expec cost (qs_NDP (map (nth (sort l)) li)),
-\end{code}
-where |sort| may be any sorting function (including |qs| itself), and where |li| is a permutation of |[0...n - 1]| such that |map (nth (sort l)) li = l| (such an |li| can easily be proven to exist).
+The right inequality is a strictly numerical affair, requiring a bit of analysis involving the harmonic series.
+As stated before, this part of the proof could be fairly directly transcribed from the paper proof, with few complications and additions, and so we will not discuss it.
+
+The left inequality is the challenging one.
+To bring it closer to the index summation, we first write |l| on the left side as |map (nth (sort l)) li|, where |sort| may be any sorting function (including |qs| itself), and where |li| is a permutation of |[0...n-1]| such that |map (nth (sort l)) li = l| (such an |li| can easily be proven to exist).
 
@@
 
+  Definition unordered_nat_pair (x y: nat): nat * nat :=
+    if x <= y then (x, y) else (y, x).
+
   Definition cmp_U (x y: nat): U comparison :=
-    ret ((if x <= y then (x, y) else (y, x)) :: nil, lookup_cmp x y).
+    ret (unordered_nat_pair x y :: nil, lookup_cmp x y).
 
-  Definition qs_U: list nat -> list nat := qs cmp_U pick_U.
+  Definition qs_U: list nat -> list nat := qs U cmp_U pick_U.
 \end{code}
-|qs_U| operates directly on a list of natural numbers representing indices into |sort l|.
+|qs_U| operates directly on lists of indices into |sort l|.
 Comparison of indices is defined by comparison of the |T| values they denote in |sort l|.
-Furthermore, rather than producing a grand total comparison count the way |NDP| does, |U| records every pair of compared indices, by using |MMT| with |Monoid_U|, the free monoid over |nat * nat| pairs, instead of the additive monoid on |nat| we used up until now.
+Furthermore, rather than producing a grand total comparison count the way |NDP| does, |U| records every pair of indices compared, by using |MMT| with |Monoid_U|, the free monoid over |nat * nat| pairs, instead of the additive monoid on |nat| we used up until now.
 
-The goal can now be rewritten using
+We now rewrite
 \begin{code}expec cost (qs_NDP (map (nth (sort l)) li))
@@
 The first equality expresses that the expected number of comparisons counted by |NDP| is equal to the expected length of the list of comparisons recorded by |U|.
-In the code, it is proved as a separate lemma by induction matching either side's |qs|'s recursion.
+In the formalization, this is a separate lemma proved by induction matching |qs|'s recursion.
 The second equality merely unfolds the definition of |monoid_expec|.
@@
 
-\parbox{\textwidth}{\begin{code}
-  forall (X: Set) (fr: X -> R) (q: list X) (t: ne_tree (list X)):
-    (forall x `In` q, expec (count x) t <= fr i) ->
-    (forall x `NotIn` q, expec (count x) t = 0) ->
-    NoDuplicates q -> expec length t <= {-"\displaystyle{\sum_{\Varid{i} \in \Varid{q}}\Varid{fr}\ \Varid{i}}"-}.
-\end{code}}
-It is trivially the case that |IJ| does not contain duplicate values.
-The second subgoal generated is\begin{code}forall (i, j) `NotIn` IJ, expec (count (i, j)) (ne_tree.map fst (qs_U li)) = 0.\end{code} Rewriting this using  the last of the identities listed for |expec| in section \ref{nondetexpec} in reverse, then rewriting the |expec| as a |monoid_expec|, and then generalizing the premise, results in the following property which is useful in its own right, as we will see later.
-\begin{code}sound_cmp_expec_0: forall i j li, (i `NotIn` li \/ j `NotIn` li) ->
-  monoid_expec (count (i, j)) (U.qs li) = 0,\end{code}
-The proof of |sound_cmp_expec_0|, which we won't show here, is by a separate induction matching |qs|'s recursion.
+\begin{code}
+  forall (X: Set) (fr: X -> R) (q: list X) (t: ne_tree (list X)),
+    (forall x `In` q, expec (count x) t <= fr x) ->
+    (forall x `NotIn` q, expec (count x) t = 0) -> expec length t <= {-"\displaystyle{\sum_{\Varid{x} \in \Varid{q}}\Varid{fr}\ \Varid{x}}"-}.
+\end{code}
+We end up with two subgoals, the first of which is \begin{code}forall (i, j) `NotIn` IJ, expec (count (i, j)) (ne_tree.map fst (qs_U li)) = 0.\end{code} Rewriting this using  the last of the identities listed for |expec| in section \ref{nondetexpec} in reverse, then rewriting the |expec| as a |monoid_expec|, and then generalizing the premise, results in
+\begin{code}
+  forall i j li, (i `NotIn` li \/ j `NotIn` li) -> monoid_expec (count (i, j)) (U.qs li) = 0,
+\end{code}
+which can be shown by induction matching |qs|'s recursion, although we will not do so in this paper. We \emph{will} use this property again in the next section.
 
-This leaves but one subgoal which, expressed with |monoid_expec|, reads
+This second subgoal, expressed with |monoid_expec|, reads
 \begin{code}
@@
 \end{code}
-This is exactly  to the property described at the beginning of this section. We prove it in the next section.
+This corresponds exactly to the property described at the beginning of this section. We prove it in the next section.
 
@@
 
-To get a stronger induction hypothesis, we slightly generalize the goal to
+Again, the proof is by induction matching |qs|'s recursion.
+But to get a better induction hypothesis, we drop the |(i, j) `In` IJ| premise (because as was shown in the last section, the statement is also true if |(i, j) `NotIn` IJ|), and add a premise saying |li| is a permutation of a contiguous sequence of indices.
 \begin{code}
-  forall i j, i < j -> forall (li: list nat) (b: nat), IndexSeq b li ->
+  forall i j, i < j -> forall (li: list nat) (b: nat), Permutation [b ... b + length li - 1] li ->
     monoid_expec (count (i, j)) (qs_U li) <= 2 / INR (S (j - i)).
 \end{code}
-The |(i, j) `In` IJ| hypothesis is dropped, because the statement is also true if |(i, j) `NotIn` IJ|, per |sound_cmp_expec_0|.
-The |IndexSeq| premise expresses that |li| is a permutation of |[b...b+length l]| (for some |b|).
-
-Again, the proof is by induction matching |qs|'s recursion. In the base case, |li| is |nil|, and the left side of the inequality reduces to 0. In the recursive case, |qs| unfolds:
+In the base case, |li| is |nil|, and the left side of the inequality reduces to 0. In the recursive case, |qs| unfolds:
 \begin{code}
 monoid_expec (count (i, j)) (
-    pi <- pick [0 ... n];
+    pi <- pick [0 ... n-1];
     let pivot := nth li pi in
     part <- partition_U pivot (remove li pi);
-    low <- qs_U (part Lt);
-    upp <- qs_U (part Gt);
-    ret (low ++ pivot :: part Eq ++ upp)
+    lower <- qs_U (part Lt); upper <- qs_U (part Gt);
+    ret (lower ++ pivot :: part Eq ++ upper)
   ) <= 2 / INR (S (j - i))
 \end{code}
-Since |cmp_U| is deterministic, |partition_U| is deterministic here as well.
-Furthermore, since we know exactly what monadic effects |partition| has in this case, we can split those effects off and then revert to simple uneffectful |filter| passes.
-Finally, the outer |monoid_expec| can be partially evaluated because of the immediately visible |pick|.
-Using these observations, the goal can be rewritten into a form that uses less monadic indirection:
-
+Since |cmp_U| is deterministic, |partition_U| is as well.
+Furthermore, since we know exactly what monadic effects |partition_U| has, we can split those effects off and revert to simple uneffectful |filter| passes.
+Finally, we rewrite using the following |monoid_expec| identity:
+\begin{code}
+  monoid_expec f (pick l >>= m) = avg (map (monoid_expec f fcmp m) l).
+\end{code}
+This way, the goal ends up in a form using less monadic indirection:
 \begin{code}
@@
     let rest := remove li pi in
-    ne_tree.map (map_fst (++ map (U.unordered_nat_pair pivot) rest)) (
-      lower <- qs_U (filter ((= Lt) fcmp lookup_cmp pivot) rest);
-      upper <- qs_U (filter ((= Gt) fcmp lookup_cmp pivot) rest);
-      ret (lower ++ (pivot :: filter ((= Eq) fcmp lookup_cmp pivot) rest) ++ upper)
-    ))) [0...n]) <= 2 / INR (S (j - i)).
+    let flt := fun c => filter ((= c) fcmp lookup_cmp pivot) rest in
+    ne_tree.map (map_fst (++ map (unordered_nat_pair pivot) rest)) (
+      lower <- qs_U (flt Lt); upper <- qs_U (flt Gt);
+      ret (lower ++ pivot :: flt Eq ++ upper)
+    ))) [0...n-1]) <= 2 / INR (S (j - i)).
 \end{code}
 Here, |map_fst| applies a function to a pair's first component.
-% todo: get this [0...n]/[0...n-1] stuff right.
 
-We now distinguish between five different cases that can occur for the nondeterministically picked |pivot|; it can either be less than |i|, equal to |i|, between |i| and |j|, equal to |j|, or greater than |j|.
+We now distinguish between five different cases that can occur for the nondeterministically picked |pivot| (which, because we are in the |U| monad, is an index).
+It can either be less than |i|, equal to |i|, between |i| and |j|, equal to |j|, or greater than |j|.
 Each case occurs a certain number of times, and has an associated expected number of |(i, j)| comparisons (coming either from the |map_fst| term representing the |partition| pass, or from the two recursive |qs_U| calls).
-To represent this split, we first rewrite the right side of the inequality to\begin{code}(2 / S (j - i) * (i - b) + 1 + 0 + 1 + 2 / S (j - i) * (b + n - j)) / S n.\end{code}
+To represent this split, we first rewrite the right side of the inequality to\[\frac{\frac{2}{S (j - i)} * (i - b) + 1 + 0 + 1 + \frac{2}{S (j - i)} * (b + n - j)}{n}\ .\]
+
 This form reflects the facts that
@@
 \end{itemize}
-With the right side of the inequality in this form, we unfold the |avg| application on the left into |sum (...) / S n|, and then cancel the division by |S n| on both sides.
+With the right side of the inequality in this form, we unfold the |avg| application on the left into |sum (...) / n|, and then cancel the division by |n| on both sides.
 Next, to actually realize the split, we apply a specialized lemma stating that
@@
    forall b i j X f n (li: list nat)
-     (g: [0..n] -> U X), IndexSeq b li ->
+     (g: [0...n-1] -> U X), Permutation [b ... b + length li - 1] li ->
     b <= i < j < b + S n -> forall ca cb, 0 <= ca -> 0 <= cb ->
@@
 \end{code}
-The proof of this lemma is somewhat tedious \todo{because..}.
-% involving a lot of repetitive arguments about permutations, etc
 
@@
       monoid_expec (count (i, j))
-        (ne_tree.map (map_fst (++ map (U.unordered_nat_pair pivot) rest)) (
+        (ne_tree.map (map_fst (++ map (unordered_nat_pair pivot) rest)) (
           foo <- qs_U (filter ((= Lt) fcmp lookup_cmp pivot) rest);
@@
   \end{code}
-Rewriting using a lemma saying that
+Since |count (i, j)| is a monoid homomorphism, we may rewrite using another lemma saying that
 \begin{code}
-  forall (A: Set) (g: m) (h: m -> nat) (t: MMT m NeTreeMonad A):
-    MonoidHomomorphism h ->
+  forall (m: Monoid) (h: m -> NatAddMonoid), monoid_homo h ->
+  forall (g: m) (A: Set) (t: MMT m NeTreeMonad A),
     monoid_expec h (ne_tree.map (map_fst (monoid_mult m g)) t) =
-    INR (h g) + monoid_expec t,
+    INR (h g) + monoid_expec h t.
 \end{code}
-the head reduces to
+This leaves
 \begin{code}
-    count (i, j) (map (U.unordered_nat_pair pivot) rest) +
+    count (i, j) (map (unordered_nat_pair pivot) rest) +
     monoid_expec (count (i, j))
@@
 From |pivot < i| and |i < j|, we have |pivot < j|.
-Since each of the comparisons in |map (U.unordered_nat_pair pivot) rest| involves the pivot element, it follows that none of them can represent comparisons between |i| and |j|.
+Since each of the comparisons in |map (unordered_nat_pair pivot) rest| involves the pivot element, it follows that none of them can represent comparisons between |i| and |j|.
 Hence, the first term vanishes.
-Furthermore, since |count (i, j)| is a monoid homomorphism, |monoid_expec_plus| lets us distribute |monoid_expec|.
+Furthermore, |monoid_expec_plus| lets us distribute |monoid_expec|.
 Since the |ret| term does not produce any comparisons either (by definition), its |monoid_expec| term vanishes, too.
@@
 \begin{code}
-monoid_expec (count (i, j))
-  (qs_U (filter ((= Lt) fcmp lookup_cmp pivot) rest)) +
-monoid_expec (count (i, j))
-  (qs_U (filter ((= Gt) fcmp lookup_cmp pivot) rest)) <= 2 / INR (S (j - i))
+monoid_expec (count (i, j)) (qs_U (filter ((= Lt) fcmp lookup_cmp pivot) rest)) +
+monoid_expec (count (i, j)) (qs_U (filter ((= Gt) fcmp lookup_cmp pivot) rest))
+  <= 2 / INR (S (j - i))
 \end{code}
-All indices in the lower filtered part denote elements that compare less than the element denoted by the pivot.
-Since the former precede the latter in |sort l|, it must be the case that the indices are less than |pivot|.
-By |sound_cmp_expec_0|, it follows that the first |qs_U| term will produce no |(i, j)| comparisons, so the first |monoid_expec| term vanishes as well, leaving
+All indices in the first filtered list denote elements that compare less than the element denoted by the pivot.
+Since the former precede the latter in |sort l|, it must be the case that these indices are all less than |pivot|.
+And since |pivot < i|, it follows that the first |qs_U| term will produce no |(i, j)| comparisons (using the property shown at the end of the previous section).
+Hence, the first |monoid_expec| term vanishes as well, leaving
 \begin{code}
@@
 \begin{itemize}
-\item If the two are equal, then |i| will not occur in the |filter term|, allowing us to invoke |sound_cmp_expec_0| yet again.
+\item If the two are equal, then |i| will not occur in the |filter term|, and so (again) no |(i, j)| comparisons are performed.
 \item If |nth (sort l) i < nth (sort l) pivot|, then we must have |i < pivot|, contradicting the assumption that |pivot < i|.
 \item If |nth (sort l) i > nth (sort l) pivot|, then we apply the induction hypothesis.
-For this, it must be shown that the filter term is a proper contiguous index sequence, which we relegated to a lemma we won't show here.
+For this, it must be shown that filtering the list of indices preserves contiguity, which follows from the fact that the indices share the order of the elements they denote in |sort l|.
 \end{itemize}
-This concludes the case where |i < pivot|. The case where |j < pivot| is symmetric. The other cases use similar arguments.
+This concludes the case where |i < pivot|. The case where |j < pivot| is symmetric. The other three cases use similar arguments.
 
@@
 
-In the interest of brevity, we have omitted innumerable details and lemmas in the description of the proof.
+In the interest of brevity, we have omitted lots of detail and various lemmas in the description of the proof.
 Still, the parts shown are reasonably faithful to the actual formalization.
@@
 \begin{itemize}
-\item We have pretended to have used ordinary natural numbers as indices into ordinary lists, completely ignoring the inescapable issues of index validity that could not be ignored in the actual formalization.
+\item In this paper we have pretended to have used ordinary natural numbers as indices into ordinary lists, completely ignoring the inescapable issues of index validity that could not be ignored in the actual formalization.
 There, we use vectors (lists whose size is part of their type) and bounded natural numbers in many places instead.
@@
 \item Using the |Program| facility to deal with quicksort's non-structural recursion was not completely as trivial as we made it out to be.
-To make proving the generated proof obligations possible, the types of |filter| and |partition| had to be sigma-decorated with modest length guarantees, because the opacity of the unspecified monad's |bind| operation prevented |Program| from being able to ``see'' the relation between the function's parameters and the arguments it passed in its recursive calls.
+Since the recursive calls were nested in lambda abstractions passed to the |bind| operation of an unspecified monad, the relation between their arguments and the function's parameters was not locally known.
+In order to be able to prove the generated proof obligations, we therefore sigma-decorated the types of |filter| and |partition| with modest length guarantees.
 \item Using the induction principles corresponding to the definitions produced by |Program| for a function already defined with two layers of monadicity quickly became unwieldy.
@@
 
-The shallow monadic embedding provides a simple but effective representation of the algorithm.
-The parameterization of the monad makes the representation very flexible, allowing a single definition to be instantiated either with basic monads (like the identity monad or bare nondeterminism monads) to get a non-instrumented version suitable for extraction and correctness proofs, or with monads that use |MMT| to add operation-counting, to support our complexity proofs.
+The formalized development successfully adopted from the original proof the idea of using a nondeterministic version of the algorithm to make the $O(n \log n)$ bound hold for any input list, the idea of taking an order-indexed perspective to reduce the problem to a sum-of-expected-comparison-counts, and the use of the standard bound for harmonic series for the strictly numerical part.
+However, for the actual reduction and the derivation of the formula for the expected comparison count, the intuitive arguments in the original proof just did not cut it.
+These parts had to essentially be reworked from scratch, building on the monadic representation of the algorithm and the various comparison counting/nondeterminism monads.
 
-Furthermore, since the shallowness of the embedding lets us re-use all standard Coq data types and facilities, including the powerful |Program Fixpoint| command, the actual algorithm definition itself is reasonably clean.
-
-The |MMT| monoid monad transformer provided the ``glue'' to add instrumentation onto a base monad.
-Its use of a monoid to represent the operation ``counter'' let us use a single set of lemmas (about |monoid_expec|) for |NDP| and |U|, as both were instances of the |M_ne_tree| monad transformed with |MMT|, differing only in the monoid used.
-Indeed, the relative ease with which we could transform the expectancy lemma into the |U| monad, a monad highly specialized for the specific order-index based counting scheme we used, further attests to the flexibility of the combination of a monadic representation and the |MMT| monad transformer.
-
-\todo{Paragraph saying that the formalized proof is faithful to the original proof, except for the part where we dropped the duplicate elements assumption.
-The purely mathematical part could be directly transcribed.
-The only theory we had to add served to support a formal treatment of the parts that the original clearly took for granted.}
-
-% old passage that may yet be useful in some form or another:
-%    Calling it an embedding almost seems misplaced, since the algorithm definitions are essentially just monadically expressed functional programs in Coq's calculus.
-Furthermore, since monads and monad transformers have been used in functional programming for decades, the basic idea should be immediately clear to any functional programmer.
-Still, it is worth noting that this pleasing convergence of subject and object languages is due solely to the fact that Coq is based on type theory.
+This shallow monadic embedding provided a simple but effective representation of the algorithm.
+Being parameterized on the monad used, it allowed a single definition to be instantiated either with basic monads (like the identity monad or bare nondeterminism monads) to get a non-instrumented version suitable for extraction and correctness proofs, or with |MMT|-transformed monads to support complexity proofs.
+Furthermore, since the shallowness of the embedding let us re-use all standard Coq data types and facilities, including the powerful |Program Fixpoint| command, the actual algorithm definition itself is reasonably clean.
 

### 2008-10-08 src/coqdeps_as_dot.hs
src: Better naming, fewer todos, some things moved around, removed cruft. Added readme, dependency graph generation, and package script.
darcs-hash:20081008020136-bab43-3cc8dea2042e8d66b1a8bac96a3dfa92269fb82d.gz

@@ -1,0 +1,38 @@
+
+import Data.List (nub, (\\))
+import Data.Map (Map)
+import qualified Data.Map as Map
+
+none :: (a -> Bool) -> [a] -> Bool
+none p = all (not . p)
+
+hide :: [String]
+hide = words "util list_utils arith_lems"
+
+main :: IO ()
+main = interact $ \s ->
+  let
+    reported_deps_map :: Map String [String]
+    reported_deps_map = Map.fromList $ map parse_line $ lines s
+
+    reported_deps :: String -> [String]
+    reported_deps n = Map.findWithDefault [] n reported_deps_map
+
+    names :: [String]
+    names = Map.keys reported_deps_map \\ hide
+
+    deep_deps_map :: Map String [String]
+    deep_deps_map = Map.fromList $ map (\n -> (n, let r = reported_deps n in r ++ concatMap deep_deps r)) names
+
+    deep_deps :: String -> [String]
+    deep_deps n = Map.findWithDefault [] n deep_deps_map
+
+    all_deps = concatMap (\n -> map (\d -> (n, d)) (deep_deps n)) names
+    depends_on x y = y `elem` deep_deps x
+    deps = [(x, y) | (x, y) <- all_deps, none (\n -> x `depends_on` n && n `depends_on` y) (names \\ [x, y])]
+  in
+    unlines $ ["digraph {"] ++ map (\(x, y) -> "\"" ++ x ++ "\" -> \"" ++ y ++ "\"") deps ++ ["}"]
+ where
+  parse_line :: String -> (String, [String])
+  parse_line l | (x, (_:xs)) <- span (/= ':') l = (strip_suffix x, map strip_suffix (tail $ words xs) \\ hide)
+  strip_suffix = takeWhile (/= '.')

### 2008-10-08 src/make_dep_graph.sh

@@ -1,0 +1,5 @@
+#!/bin/sh
+ghc -XPatternGuards --make coqdeps_as_dot.hs &&
+  coqdep -I . *.v | ./coqdeps_as_dot | dot -Nfontsize=12 -Tpng > dep_graph.png
+
+# We can't just use runhaskell, because it outputs readline crap that dot chokes on...

### 2008-10-08 src/package.sh

@@ -1,0 +1,7 @@
+#!/bin/sh
+set -e
+mkdir qs-avg
+./make_dep_graph.sh
+cp *.v README.html SConstruct dep_graph.png qs-avg
+tar -czf ../qs-avg-`date +"%Y%m%dT%H%M"`.tar.gz qs-avg
+rm -rf qs-avg

### 2008-10-08 paper/paper.tex
bibplusauthorpatch
I ruthlessly tidied the .bib entries. I added a \thanks to the authors. I changed their
priority order.

darcs-hash:20080926103109-d98fc-470b9a02cd967da868ada0361f464257ac115514.gz
paper: Various small fixes, as discussed Friday 2008-09-26.
darcs-hash:20081008030956-bab43-8b3175aaad991e7488d883ef94b324977196e119.gz

@@
 
-\newcommand{\theauthors}{James McKinna and Eelis van der Weegen}
+\newcommand{\eelis}{Eelis van der Weegen}
+\newcommand{\james}{James McKinna}
+
+\newcommand{\theauthors}{\eelis\ and \james}
 \newcommand{\thetitle}{A Machine-checked Proof of the Average-case Complexity of Quicksort in Coq}
@@
 
-\author{\theauthors}
+\author{\eelis\thanks{Research carried out by the first author as part of the Master's programme in ``Foundations''} and \james \\
+eelis@@eelis.net, james.mckinna@@cs.ru.nl}
 
-\institute{Department of Computer Science, University of Nijmegen, the Netherlands}
+\institute{Department of Computer Science, University of Nijmegen \\
+P.O. Box 9010, 6500 GL Nijmegen, The Netherlands}
 
@@
 The first part shows that the total number of comparisons performed by the algorithm (the usual complexity metric for sorting algorithms) can be written as a sum of expected comparison counts for individual pairs of input list elements.
-The second part derives from the algorithm a specific formula for this expectancy.
-The third and last part employs some analysis involving the harmonic series to derive the $O(n \log n)$ bound from the sum-of-expectancies.
+The second part derives from the algorithm a specific formula for this expectation.
+The third and last part employs some analysis involving the harmonic series to derive the $O(n \log n)$ bound from the sum-of-expectations.
 
@@
 
+A completely different approach to type-theoretic analysis of computational complexity is to devise a special-purpose type theory in which the types of terms include some form of complexity guarantees.
+This is the approach taken in \cite{constable}, for example.
+However, as far as we can see, this method is not suitable for the formalization of a traditional complexity proof in which an algorithm is first defined and then separately analyzed.
+We are not aware of practical techniques usable for this purpose apart from the one we present in this paper.
+
 \section{A Shallow Monadic Embedding}
@@
 We make those uses observable by writing the algorithm as a monadically expressed function, parameterizing it on both the monad itself and on the comparison operation.
-Expressed this way, a deterministic quicksort that simply selects the head of the input list as its pivot element, and uses two simple filter passes to partition the input list, looks as follows:
+Expressed this way, a deterministic Quicksort that simply selects the head of the input list as its pivot element, and uses two simple filter passes to partition the input list, looks as follows:
 
@@
 For a general introduction to monadic programming and monad laws, see \cite{wadler93monads}.
-The notation ``|x <- y; z|'', shorthand for |y >>= \x => z| (where |x| may occur in |z|) is known as ``do-notation''.
+Furthermore, we use Haskell's ``do-notation'', and freely use standard monadic functions such as |liftM| and |filterM|.
 
-We use Coq's |Program| facility \cite{subsetcoercions} to cope with the fact that Quicksort's recursion is non-structural, by using the input list length as a measure.
-The proof obligations generated by |Program| for the above definition and measure are trivial enough for Coq to prove mostly by itself.
+We use Coq's |Program Fixpoint| facility \cite{subsetcoercions} to cope with Quicksort's non-structural recursion, specifying list length as an input measure function that is separately shown to strongly decrease for each recursive call.
+For |qs|, these proof obligations are trivial enough for Coq to prove mostly by itself.
+
+For recursive functions defined this way, Coq does not automatically define corresponding induction principles matching the recursive call structure.
+Hence, for this |qs| definition as well as the one we will introduce in section \ref{nondetexpec}, we have defined these induction principles manually.
+To make their use as convenient as possible, we further customized and specialized them to take advantage of specific monad properties.
+We will ignore these issues in this paper, and will henceforth simply say: ``by induction on |qs|, ...''.
 
@@
 
-For reasons that will become clear in later sections, we compose the monad we will instantiate the above definitions with using a monad transformer |MMT| (for ``monoid monad transformer''), which piggybacks a monoid onto an existing monad by pairing.
+For reasons that will become clear in later sections, we compose the monad we will instantiate the above definitions with using a monad transformer \cite{monadtrans} |MMT| (for ``monoid monad transformer''), which piggybacks a monoid onto an existing monad by pairing.
 
@@
 \begin{proof}
-  The proof is by induction matching |qs|'s recursion.
-For an empty input list, we have |cost (qs_SP nil) = cost (ret nil) = 0 <= sqrd (length l)|.
-For a non-empty input list |h :: t|, the cost decomposes into
+  The proof is by induction on |qs|. For an empty input list, we have |cost (qs_SP nil) = cost (ret nil) = 0 <= sqrd (length l)|. For a non-empty input list |h :: t|, the cost decomposes into
   \begin{code}
@@
 Instead of a two-way comparison operation producing a monadic |bool|, it takes a three-way comparison operation producing a monadic |comparison|, which is an enumeration with values |Lt|, |Eq|, and |Gt|.
+We represent the resulting partitioning by a function of type |comparison -> list T| rather than a record or tuple type containing three lists, because in the actual source code, this saves us from having to constantly map |comparison| values to corresponding record field accessors or tuple projections.
 
@@
 
-The last thing needed before the main theorem can be stated, is the notion of big-O complexity. We use a standard textbook definition, except that we incorporate a measure function parameter:
+The last thing needed before the main theorem can be stated, is the notion of big-O complexity. We use a standard textbook definition, except that we incorporate a measure function parameter |m|:
 
@@
 \begin{code}
-  forall l, expec cost (qs_NDP l) <= 2 * length l * S (log2ceil (length l)),
+  forall l, expec cost (qs_NDP l) <= 2 * length l * (1 + log2ceil (length l)),
 \end{code}
@@
 More specifically, if $X \equiv X_{I_0} \ldots X_{I_{n-1}}$ is the input list, with $I$ a permutation of |[0 ...
-n-1]| such that $X_0 \ldots X_n$ is sorted, then the expected comparison count for any $X_i$ and $X_j$ with $i < j$ is at most $2 / S (j - i)$.
+n-1]| such that $X_0 \ldots X_n$ is sorted, then the expected comparison count for any $X_i$ and $X_j$ with $i < j$ is at most $2 / (1 + j - i)$.
 In other words, the expected comparison count for two input list elements is bounded by a simple function of the number of list elements that separate the two in the sort order.
@@
 \begin{code}
-  expec cost (qs_NDP l) <= {-"\displaystyle{\sum_{(i, j) \in \Varid{IJ}}\frac{2}{S (j - i)}}"-} <= 2 * length l * S (log2ceil (length l)),
+  expec cost (qs_NDP l) <= {-"\displaystyle{\sum_{(i, j) \in \Varid{IJ}}}"-} ecc i j <= 2 * length l * (1 + log2ceil (length l)),
 \end{code}
-where $\Varid{IJ} := \{ (i, j) \in [0, |length l|) \mid i < j \}$.
+where $\Varid{IJ} := \{ (i, j) \in [0, |length l|) \mid i < j \}$, and |ecc i j := 2 / (1 + j - i)|.
 
@@
 The first equality expresses that the expected number of comparisons counted by |NDP| is equal to the expected length of the list of comparisons recorded by |U|.
-In the formalization, this is a separate lemma proved by induction matching |qs|'s recursion.
+In the formalization, this is a separate lemma proved by induction on |qs|.
 The second equality merely unfolds the definition of |monoid_expec|.
@@
 \begin{code}
-  expec length (ne_tree.map fst (qs_U li)) <= {-"\displaystyle{\sum_{(i, j) \in \Varid{IJ}}\frac{2}{S (j - i)}}"-}.
+  expec length (ne_tree.map fst (qs_U li)) <= {-"\displaystyle{\sum_{(i, j) \in \Varid{IJ}}}"-} ecc i j.
 \end{code}
@@
 \end{code}
-which can be shown by induction matching |qs|'s recursion, although we will not do so in this paper. We \emph{will} use this property again in the next section.
+which can be shown by induction on |qs|, although we will not do so in this paper. We \emph{will} use this property again in the next section.
 
-This second subgoal, expressed with |monoid_expec|, reads
+The second subgoal, expressed with |monoid_expec|, reads
 \begin{code}
-  forall (i, j) `In` IJ, monoid_expec (count (i, j)) (qs_U li) <= 2 / INR (S (j - i)).
+  forall (i, j) `In` IJ, monoid_expec (count (i, j)) (qs_U li) <= ecc i j.
 \end{code}
@@
 
-Again, the proof is by induction matching |qs|'s recursion.
+Again, the proof is by induction on |qs|.
 But to get a better induction hypothesis, we drop the |(i, j) `In` IJ| premise (because as was shown in the last section, the statement is also true if |(i, j) `NotIn` IJ|), and add a premise saying |li| is a permutation of a contiguous sequence of indices.
@@
   forall i j, i < j -> forall (li: list nat) (b: nat), Permutation [b ... b + length li - 1] li ->
-    monoid_expec (count (i, j)) (qs_U li) <= 2 / INR (S (j - i)).
+    monoid_expec (count (i, j)) (qs_U li) <= ecc i j.
 \end{code}
@@
     ret (lower ++ pivot :: part Eq ++ upper)
-  ) <= 2 / INR (S (j - i))
+  ) <= ecc i j.
 \end{code}
@@
       ret (lower ++ pivot :: flt Eq ++ upper)
-    ))) [0...n-1]) <= 2 / INR (S (j - i)).
+    ))) [0...n-1]) <= ecc i j.
 \end{code}
@@
 Each case occurs a certain number of times, and has an associated expected number of |(i, j)| comparisons (coming either from the |map_fst| term representing the |partition| pass, or from the two recursive |qs_U| calls).
-To represent this split, we first rewrite the right side of the inequality to\[\frac{\frac{2}{S (j - i)} * (i - b) + 1 + 0 + 1 + \frac{2}{S (j - i)} * (b + n - j)}{n}\ .\]
-
+To represent this split, we first rewrite the right side of the inequality to\[\frac{\Varid{ecc}\ i\ j * (i - b) + 1 + 0 + 1 + \Varid{ecc}\ i\ j * (b + n - j)}{n}\ .\]
 This form reflects the facts that
 \begin{itemize}
-\item the case where |pivot| is less than |i| occurs |i - b| times, and in each instance, the expected number of |(i, j)| comparisons is no more than |2 / S (j - i)|;
+\item the case where |pivot| is less than |i| occurs |i - b| times, and in each instance, the expected number of |(i, j)| comparisons is no more than |ecc i j|;
 \item the case where the |pivot| is equal to |i| occurs once, and in this case no more than a single |(i, j)| comparison is expected;
@@
 \item the case where the |pivot| is equal to |j| occurs once, and in this case no more than a single |(i, j)| comparison is expected;
-\item the case where the |pivot| is greater than |j| occurs |b + n - j| times, and in each instance, the expected number of |(i, j)| comparisons is no more than |2 / S (j - i)|.
+\item the case where the |pivot| is greater than |j| occurs |b + n - j| times, and in each instance, the expected number of |(i, j)| comparisons is no more than |ecc i j|.
 \end{itemize}
@@
           ret (foo ++ (pivot :: filter ((= Gt) fcmp lookup_cmp pivot) rest) ++ bar)))
-        <= 2 / S (j - i).
+        <= ecc i j.
   \end{code}
@@
       ret  (foo ++ (nth v pi :: filter ((= Eq) fcmp lookup_cmp pivot) rest) ++ bar))
-    <= 2 / S (j - i).
+    <= ecc i j.
 \end{code}
@@
 Hence, the first term vanishes.
-Furthermore, |monoid_expec_plus| lets us distribute |monoid_expec|.
+Furthermore, |monoid_expec_plus| lets us distribute |monoid_expec| over the |bind| applications.
 Since the |ret| term does not produce any comparisons either (by definition), its |monoid_expec| term vanishes, too.
@@
 monoid_expec (count (i, j)) (qs_U (filter ((= Gt) fcmp lookup_cmp pivot) rest))
-  <= 2 / INR (S (j - i))
+  <= ecc i j.
 \end{code}
-All indices in the first filtered list denote elements that compare less than the element denoted by the pivot.
+All indices in the first filtered list denote elements that are less than the element denoted by the pivot.
 Since the former precede the latter in |sort l|, it must be the case that these indices are all less than |pivot|.
@@
 monoid_expec (count (i, j))
-  (qs_U (filter ((= Gt) fcmp lookup_cmp pivot) rest)) <= 2 / INR (S (j - i)).
+  (qs_U (filter ((= Gt) fcmp lookup_cmp pivot) rest)) <= ecc i j.
 \end{code}
@@
 In the interest of brevity, we have omitted lots of detail and various lemmas in the description of the proof.
-Still, the parts shown are reasonably faithful to the actual formalization.
-For completeness, we briefly list some complicating factors that we have ignored.
-In some cases, resolving these took significant effort.
+Still, the parts shown are reasonably faithful to the actual formalization, with two notable exceptions.
 
-\begin{itemize}
-\item In this paper we have pretended to have used ordinary natural numbers as indices into ordinary lists, completely ignoring the inescapable issues of index validity that could not be ignored in the actual formalization.
+First, we have pretended to have used ordinary natural numbers as indices into ordinary lists, completely ignoring issues of index validity that could not be ignored in the actual formalization.
 There, we use vectors (lists whose size is part of their type) and bounded natural numbers in many places instead.
 Using these substantially reduces the amount of |i < length l| proofs that need to be produced, converted, and passed around, but this solution is still far from painless.
-\item Using the |Program| facility to deal with quicksort's non-structural recursion was not completely as trivial as we made it out to be.
-Since the recursive calls were nested in lambda abstractions passed to the |bind| operation of an unspecified monad, the relation between their arguments and the function's parameters was not locally known.
-In order to be able to prove the generated proof obligations, we therefore sigma-decorated the types of |filter| and |partition| with modest length guarantees.
-\item Using the induction principles corresponding to the definitions produced by |Program| for a function already defined with two layers of monadicity quickly became unwieldy.
+
+Second, using the |Program| facility to deal with Quicksort's non-structural recursion is not completely as trivial as we made it out to be.
+Since the recursive calls are nested in lambda abstractions passed to the |bind| operation of an unspecified monad, the relation between their arguments and the function's parameters is not locally known, resulting in unprovable proof obligations.
+To make these provable, we $\Sigma$-decorated the types of |filter| and |partition| in the actual formalization with modest length guarantees.
+% \item Using the induction principles corresponding to the definitions produced by |Program| for a function already defined with two layers of monadicity quickly became unwieldy.
 To deal with this, we defined a series of specialized induction principles for specific instances of |qs|, which put the recursive proof obligation in as simple terms as possible, with as little monadic indirection as possible, using among other things the observations regarding determinism of the comparison operation, as described in section \ref{finishing}.
-\end{itemize}
 
@@
 The formalized development successfully adopted from the original proof the idea of using a nondeterministic version of the algorithm to make the $O(n \log n)$ bound hold for any input list, the idea of taking an order-indexed perspective to reduce the problem to a sum-of-expected-comparison-counts, and the use of the standard bound for harmonic series for the strictly numerical part.
-However, for the actual reduction and the derivation of the formula for the expected comparison count, the intuitive arguments in the original proof just did not cut it.
-These parts had to essentially be reworked from scratch, building on the monadic representation of the algorithm and the various comparison counting/nondeterminism monads.
+However, for the actual reduction and the derivation of the formula for the expected comparison count, the intuitive arguments essentially had to be reworked from scratch, building on the monadic representation of the algorithm and the various comparison counting/nondeterminism monads.
 
-This shallow monadic embedding provided a simple but effective representation of the algorithm.
-Being parameterized on the monad used, it allowed a single definition to be instantiated either with basic monads (like the identity monad or bare nondeterminism monads) to get a non-instrumented version suitable for extraction and correctness proofs, or with |MMT|-transformed monads to support complexity proofs.
-Furthermore, since the shallowness of the embedding let us re-use all standard Coq data types and facilities, including the powerful |Program Fixpoint| command, the actual algorithm definition itself is reasonably clean.
+The shallow monadic embedding provides a simple but effective representation of the algorithm.
+Being parameterized on the monad used, it allows a single definition to be instantiated either with basic monads (like the identity monad or bare nondeterminism monads) to get a non-instrumented version suitable for extraction and correctness proofs, or with |MMT|-transformed monads to support complexity proofs.
+Furthermore, since the shallowness of the embedding lets us re-use all standard Coq data types and facilities, including the powerful |Program Fixpoint| command, the actual algorithm definition itself is reasonably clean.
 
-For future work, an interesting next challenge is to try and formalize a (likely more complex) proof that the $O(n \log n)$ bound holds also for deterministic quicksorts (with sensible pivot selection strategies).
+For future work, an interesting next challenge is to try and formalize a (likely more complex) proof that the $O(n \log n)$ bound holds also for deterministic Quicksorts (with sensible pivot selection strategies).
 

### 2008-10-10 paper/paper.tex
final-submission-tweaks
darcs-hash:20081010192138-d98fc-9330e20208eddd813e28da7a4736fe7b73c0db32.gz

@@
 
-Defining |cost| and |result| as the first and second projection, respectively, we trivially have identities such as |cost (ret_SP x) = 0|, |cost (x `infixbind_SP` f) = cost x + cost (f (result x))|, and |cost (le_SP x y) = 1|.
+Defining |cost| and |result| as the first and second projection, respectively, we trivially have identities such as |cost (ret_SP x) = 0|, |cost (le_SP x y) = 1|, and |cost (x `infixbind_SP` f) = cost x + cost (f (result x))|.
 This very modest amount of machinery is sufficient for a straightforward proof of Quicksort's quadratic worst-case complexity.
 
-\begin{propo}|qs_worst: forall l, cost (qs_SP l) <= sqrd (length l).|\footnote{(We do not use big-O notation for this simple statement, as it would only obfuscate.
-Big-O complexity is discussed in section \ref{statement}.)}
+\begin{propo}|qs_worst: forall l, cost (qs_SP l) <= sqrd (length l).|\footnote{We do not use big-O notation for this simple statement, as it would only obfuscate.
+Big-O complexity is discussed in section \ref{statement}.}
 \end{propo}
@@
 \begin{proof}
-  The proof is by induction on |qs|. For an empty input list, we have |cost (qs_SP nil) = cost (ret nil) = 0 <= sqrd (length l)|. For a non-empty input list |h :: t|, the cost decomposes into
+  The proof is by induction on |qs|. For |l==nil|, we have {|cost (qs_SP nil) = cost (ret nil) = 0 <= sqrd (length l)|}. For |l==h :: t|, the cost decomposes into
   \begin{code}
@@
 
-We now expand the technique in preparation of the average-case complexity proof.
+We now expand the technique in preparation for the average-case complexity proof.
 
@@
 Next, we redefine |qs| to use |partition|, and have it take as an additional parameter a |pick| operation, representing nondeterministic selection of an element of a non-empty list of choices.
-(An |ne_list T| is a non-empty list of |T|'s.)
+An |ne_list T| is a non-empty list of |T|'s, inductively defined in the obvious way.
 
@@
 
-Using tree nondeterminism instead solves the problem.
+Using tree nondeterminism instead solves the problem: we introduce the type |ne_tree| of non-empty trees, building on |ne_list|: 
 \begin{code}
@@
 
-In the interest of brevity, we have omitted lots of detail and various lemmas in the description of the proof.
+In the interest of brevity, we have omitted lots of gory details and various lemmas in the description of the proof.
 Still, the parts shown are reasonably faithful to the actual formalization, with two notable exceptions.
@@
 
-For all the gruesome details, we refer the interested reader to the Coq source files.
-
 The formalized development successfully adopted from the original proof the idea of using a nondeterministic version of the algorithm to make the $O(n \log n)$ bound hold for any input list, the idea of taking an order-indexed perspective to reduce the problem to a sum-of-expected-comparison-counts, and the use of the standard bound for harmonic series for the strictly numerical part.
@@
 
-For future work, an interesting next challenge is to try and formalize a (likely more complex) proof that the $O(n \log n)$ bound holds also for deterministic Quicksorts (with sensible pivot selection strategies).
+We have shown that it is straightforward to give a fully formal treatment in type theory of a classical result in complexity theory.
+This clearly shows the utility and applicability of the general monadic framework for worst- and average-case complexity that we have developed.
 

### 2009-01-27 paper/paper.tex
paper: Straightforward fixes suggested by referees.
darcs-hash:20090127222406-bab43-b095ab766e5dc20b9eb16d010f21e8c844b1d67c.gz

@@
 \usepackage{color}
+\usepackage{amsmath}
 \usepackage[colorlinks,pdftitle=\thetitle,pdfauthor=\theauthors]{hyperref}
@@
 For a general introduction to monadic programming and monad laws, see \cite{wadler93monads}.
-Furthermore, we use Haskell's ``do-notation'', and freely use standard monadic functions such as |liftM| and |filterM|.
+Furthermore, we use Haskell's ``do-notation'', and freely use standard monadic functions such as |liftM| (of type |forall (M: Monad) (A B: Set), (A -> B) -> (M A -> M B)|) and |filterM| (of type |forall (M: Monad) (A: Set), (A -> M bool) -> list A -> M (list A)|).
 
@@
   \begin{code}
-    cost (filter (le h) t) + cost (qs_SP (result (filter (lt h) t))) +
+    cost (filter (le h) t) + cost (qs_SP (result (filter (le h) t))) +
     cost (filter (gt h) t) + cost (qs_SP (result (filter (gt h) t))) +
-    cost (ret (result (qs_SP (result (filter (lt h) t))) ++
+    cost (ret (result (qs_SP (result (filter (le h) t))) ++
      h :: result (qs_SP (result (filter (gt h) t))))).
@@
 Instead of a two-way comparison operation producing a monadic |bool|, it takes a three-way comparison operation producing a monadic |comparison|, which is an enumeration with values |Lt|, |Eq|, and |Gt|.
-We represent the resulting partitioning by a function of type |comparison -> list T| rather than a record or tuple type containing three lists, because in the actual source code, this saves us from having to constantly map |comparison| values to corresponding record field accessors or tuple projections.
+We represent the resulting partitioning by a function of type |comparison -> list T| rather than a record or tuple type containing three lists, because in the actual formalization, this saves us from having to constantly map |comparison| values to corresponding record field accessors or tuple projections.
+This is only a matter of minor convenience; a record or tuple could have been used instead without problems.
 
@@
 The functions |nth| and |remove| select and remove the $n$th element of a list, respectively.
+
+Note that the Quicksort definition in section \ref{embed} could have been implemented with a |partition| pass as well, which might well have made the worst-case proof even simpler.
+We chose not to do this, in order to emphasize that the properties the average-case proof demands of the algorithm rule out the naive but familiar implementation using |filter| passes.
 
@@
 
-|expec| gives rise to a host of identities, such as
+The function |expec| gives rise to a host of identities, such as
 
-\begin{code}
-  0 <= expec f t
-  expec (fun x => f x + g x) t = expec f t + expec g t.
-  expec ((* c) fcmp f) = (* c) fcmp expec f
-
-  (forall x `In` t -> f x <= g x) -> expec f t <= expec g t.
-  (forall x `In` t -> f x = c) -> expec f t = c
-  (forall x, x `In` t -> f x = 0) <-> expec f t = 0
-
-  expec f (t >>= (ret fcmp g)) = expec (f fcmp g) t = expec f (ne_tree.map g t)
-\end{code}
+\begin{align}
+  0 &\le |expec f t| \nonumber \\
+  |expec (fun x => f x + g x) t| &= |expec f t + expec g t| \nonumber \\
+  |expec ((* c) fcmp f)| &= |(* c) fcmp expec f| \nonumber \\
+  |(forall x `In` t -> f x <= g x) -> expec f t| &\le |expec g t| \nonumber \\
+  |(forall x `In` t -> f x = c) -> expec f t| &= c \nonumber \\
+  |(forall x `In` t -> f x = 0) <-> expec f t| &= 0 \nonumber \\
+  |expec f (t >>= (ret fcmp g))| &= |expec (f fcmp g) t| \nonumber \\
+    &= |expec f (ne_tree.map g t)| \label{expec_bind_lem}
+\end{align}
 
@@
 
-  Definition cmp_NDP (x y: T): M bool := ret_ne_tree (1, cmp x y).
+  Definition cmp_NDP (x y: T): M_NDP bool := ret_ne_tree (1, cmp x y).
 
@@
 Thanks to the property discussed at the start of the previous section, % perhaps too far back
-|qs_avg| follows as a corollary from the stronger statement that
+|qs_avg| follows as a corollary from the stronger statement
 \begin{code}
-  forall l, expec cost (qs_NDP l) <= 2 * length l * (1 + log2ceil (length l)),
+  qs_expec_cost: forall l, expec cost (qs_NDP l) <= 2 * length l * (1 + log2ceil (length l)),
 \end{code}
@@
 
-After rewriting with the last of the identities listed for |expec| in section \ref{nondetexpec}, the goal becomes
+After rewriting with identity \ref{expec_bind_lem} from section \ref{nondetexpec}, the goal becomes
 \begin{code}
@@
 
-In the interest of brevity, we have omitted lots of gory details and various lemmas in the description of the proof.
+In the interest of brevity, we have omitted lots of detail and various lemmas in the description of the proof.
 Still, the parts shown are reasonably faithful to the actual formalization, with two notable exceptions.

### 2009-02-10 paper/paper.tex
paper: Trivial formatting fix.
Ignore-this: 4047beb586aad32da542c5519ade3223

darcs-hash:20090210114049-bab43-e6c487754c544b98927d0550ff6f1ac32a34b403.gz
paper: Nicer Cartesian product symbol.
Ignore-this: 726119862465770478e7aaa1c4bc1a2e

darcs-hash:20090210114107-bab43-340788204d9218ea8795a3351b72a5aa283db47e.gz

@@
 
+%format prod = "\hspace{-1mm}\times\hspace{-1mm} "
 %format sp = "\ "
@@
 
-  Let C_MMT (T: Set): Set := monad (monoid * T).
+  Let C_MMT (T: Set): Set := monad (monoid prod T).
 
@@
 \begin{code}
-  Definition Monoid_U: Monoid := (list (nat * nat), nil, ++).
+  Definition Monoid_U: Monoid := (list (nat prod nat), nil, ++).
 
@@
 
-  Definition unordered_nat_pair (x y: nat): nat * nat :=
+  Definition unordered_nat_pair (x y: nat): nat prod nat :=
     if x <= y then (x, y) else (y, x).
@@
 \begin{itemize}
-\item If the two are equal, then |i| will not occur in the |filter term|, and so (again) no |(i, j)| comparisons are performed.
+\item If the two are equal, then |i| will not occur in the |filter| term, and so (again) no |(i, j)| comparisons are performed.
 \item If |nth (sort l) i < nth (sort l) pivot|, then we must have |i < pivot|, contradicting the assumption that |pivot < i|.

### 2009-02-10 paper/paper.tex
paper: Fix awkward code layout in M_ne_tree definition.
Ignore-this: 305cffecce046ade027e39effdb3533f

darcs-hash:20090210133833-bab43-dd4d4e0c5fa9d52d579fbcc8faa16639afad69f6.gz
paper: Make some formulas slightly less cramped by putting recursive qs calls on separate lines.
Ignore-this: ede624d932bb56b09857f59561f4a7aa

darcs-hash:20090210142447-bab43-5a518aad267ccce815c17512d101547dabefe1b5.gz
paper: Add paragraph about existing work; address some other referee comments.
Ignore-this: fd7b93d1630b9b83a22495ffdcf67daa

darcs-hash:20090210142617-bab43-447e2bf27f361e2576ceac342a8023fb390b2dd2.gz

@@
 
+\paragraph{Related work}
+
+In his Ph.D thesis \cite{hurd}, Hurd presents an approach to formal analysis of probabilistic programs based on a comprehensive formalization of measure-theoretic constructions of probability spaces, representing probabilistic programs using a state-transforming monad in which bits from an infinite supply of random bits may be consumed.
+In \cite{mohringaudebaud}, Audebaud and Paulin-Mohring describe a different monadic approach in which programs are interpreted directly as measures representing probability distributions.
+A set of axiomatic rules is defined for estimating the probability that programs interpreted this way satisfy certain properties.
+
+Compared to these approaches, our infrastructure for reasoning about nondeterministic programs is far less ambitious, supporting only a very simplistic notion of expectation, computed simply by monadically collecting all possible outputs (which we only permit to be finitely many), and then taking the average of a measure applied to these outputs.
+In particular, we do not concern ourselves at all with probability distributions.
+
 A completely different approach to type-theoretic analysis of computational complexity is to devise a special-purpose type theory in which the types of terms include some form of complexity guarantees.
 This is the approach taken in \cite{constable}, for example.
-However, as far as we can see, this method is not suitable for the formalization of a traditional complexity proof in which an algorithm is first defined and then separately analyzed.
-We are not aware of practical techniques usable for this purpose apart from the one we present in this paper.
 
@@
 Combined, these two traits ensure that the $O(n \log n)$ average-case bound holds not just averaged over all possible input lists, but for each individual input list as well.
-This greatly simplifies things, because it means that if we prove the bound for an arbitrary input, the global bound immediately follows.
+This greatly simplifies things, because it removes the need to carefully track input distributions in order to show that nice inputs (for which the original deterministic version of the algorithm performs well) sufficiently outnumber bad inputs (for which the original deterministic version of the algorithm performs poorly).
+In particular, it means that once we prove that the bound holds for an arbitrary input, the global bound immediately follows.
+It also means that for a key lemma near the end of our proof, we can use straightforward induction over the algorithm's recursive call structure, without having to show that given appropriately distributed inputs, the partition step yields lists that are again appropriately distributed.
+The latter is a typical concern in more ambitious approaches for average-case complexity analysis \cite[for example]{modularcalculus} and analysis of probabilistic algorithms.
 
+While we sympathize with the view that changing the algorithm under investigation amounts to cheating, we will nevertheless simply follow the textbook proof (which represents common practice) in this regard.
 Hence, we now modify our |qs| definition to use a three-way partition pass, and to use nondeterministic pivot selection.
@@
 
-Note that the Quicksort definition in section \ref{embed} could have been implemented with a |partition| pass as well, which might well have made the worst-case proof even simpler.
+Note that the deterministic Quicksort definition in section \ref{embed} could have been implemented with a |partition| pass as well, which might well have made the worst-case proof even simpler.
 We chose not to do this, in order to emphasize that the properties the average-case proof demands of the algorithm rule out the naive but familiar implementation using |filter| passes.
@@
 
-  Fixpoint bind_ne_tree (A B: Set) (m: ne_tree A) (k: A -> ne_tree B): ne_tree B :=
+  Fixpoint bind_ne_tree (A B: Set)
+    (m: ne_tree A) (k: A -> ne_tree B): ne_tree B :=
       match m with
@@
       end.
-\end{code}
-\begin{code}
+
   Definition M_ne_tree: Monad := Build_Monad ne_tree bind_ne_tree ret_ne_tree.
@@
 Comparison of indices is defined by comparison of the |T| values they denote in |sort l|.
-Furthermore, rather than producing a grand total comparison count the way |NDP| does, |U| records every pair of indices compared, by using |MMT| with |Monoid_U|, the free monoid over |nat * nat| pairs, instead of the additive monoid on |nat| we used up until now.
+Furthermore, rather than producing a grand total comparison count the way |NDP| does, |U| records every pair of indices compared, by using |MMT| with |Monoid_U|, the free monoid over |nat prod nat| pairs, instead of the additive monoid on |nat| we used up until now.
 
@@
 \end{code}
-We end up with two subgoals, the first of which is \begin{code}forall (i, j) `NotIn` IJ, expec (count (i, j)) (ne_tree.map fst (qs_U li)) = 0.\end{code} Rewriting this using  the last of the identities listed for |expec| in section \ref{nondetexpec} in reverse, then rewriting the |expec| as a |monoid_expec|, and then generalizing the premise, results in
+We end up with two subgoals, the first of which is \begin{code}forall (i, j) `NotIn` IJ, expec (count (i, j)) (ne_tree.map fst (qs_U li)) = 0.\end{code} Rewriting this using identity \ref{expec_bind_lem} from section \ref{nondetexpec} in reverse, then rewriting the |expec| as a |monoid_expec|, and then generalizing the premise, results in
 \begin{code}
@@
     part <- partition_U pivot (remove li pi);
-    lower <- qs_U (part Lt); upper <- qs_U (part Gt);
+    lower <- qs_U (part Lt);
+    upper <- qs_U (part Gt);
     ret (lower ++ pivot :: part Eq ++ upper)
@@
     ne_tree.map (map_fst (++ map (unordered_nat_pair pivot) rest)) (
-      lower <- qs_U (flt Lt); upper <- qs_U (flt Gt);
+      lower <- qs_U (flt Lt);
+      upper <- qs_U (flt Gt);
       ret (lower ++ pivot :: flt Eq ++ upper)

### 2009-02-18 paper/paper.tex
Final tweaks (hopefully).
Ignore-this: b5ffc6e4392932040dd60aed83516b81

darcs-hash:20090218164704-bab43-1637651ede1cac2f3cafdd5e3420210817447d7e.gz

@@
 
-The formalization is based on the ``paper proof'' in \cite{introtoalgos}, which consists of three parts.
+The formalisation is based on the ``paper proof'' in \cite{introtoalgos}, which consists of three parts.
 The first part shows that the total number of comparisons performed by the algorithm (the usual complexity metric for sorting algorithms) can be written as a sum of expected comparison counts for individual pairs of input list elements.
@@
 
-The next two sections detail the actual formalized proof.
+The next two sections detail the actual formalised proof.
 Section \ref{reduction} corresponds to the first part in the original proof described above, and shows how the main theorem can be split up into a lemma (stated in terms of another specialized monad) giving a formula for the expected comparison count for individual pairs of input element, and a strictly numerical part.
@@
 Section \ref{finishing} finishes the proof by proving the lemma about the expected comparison count for individual input list elements.
-Since this is the part where the original proof omits the most detail, and makes the assumption regarding duplicate elements, and where we really have to reason in detail about the behavior of the algorithm, it is by far the most involved part of the formalization.
+Since this is the part where the original proof omits the most detail, and makes the assumption regarding duplicate elements, and where we really have to reason in detail about the behavior of the algorithm, it is by far the most involved part of the formalisation.
 
@@
 
-The Coq source files containing the entire formalization can be downloaded from \url{http://www.eelis.net/research/quicksort/}. The Coq version used is 8.2 beta 2.
+The Coq source files containing the entire formalisation can be downloaded from \url{http://www.eelis.net/research/quicksort/}. The Coq version used is 8.2 beta 2.
 
@@
 
-In his Ph.D thesis \cite{hurd}, Hurd presents an approach to formal analysis of probabilistic programs based on a comprehensive formalization of measure-theoretic constructions of probability spaces, representing probabilistic programs using a state-transforming monad in which bits from an infinite supply of random bits may be consumed.
+In his Ph.D thesis \cite{hurd}, Hurd presents an approach to formal analysis
+of probabilistic programs based on a comprehensive formalisation of
+measure-theoretic constructions of probability spaces, representing
+probabilistic programs using a state-transforming monad in which bits from an
+infinite supply of random bits may be consumed. He even mentions the problem
+of proving the average-case complexity of Quicksort, but leaves it for future work.
+
 In \cite{mohringaudebaud}, Audebaud and Paulin-Mohring describe a different monadic approach in which programs are interpreted directly as measures representing probability distributions.
@@
 
-Compared to these approaches, our infrastructure for reasoning about nondeterministic programs is far less ambitious, supporting only a very simplistic notion of expectation, computed simply by monadically collecting all possible outputs (which we only permit to be finitely many), and then taking the average of a measure applied to these outputs.
-In particular, we do not concern ourselves at all with probability distributions.
+Compared to these approaches, our infrastructure for reasoning about nondeterministic programs is rather less ambitious, in that we only consider finite expectation based on na{\"\i}ve counting probability, using a monad for nondeterminism which correctly supports weighted expectation.
+In particular, we do not need to reason explicitly with probability distributions.
 
@@
 
-Since we are working in Coq, we already have at our disposal a full functional programming language, in the form of Coq's pCIC \cite{coq'art}.
-However, just writing the algorithm as an ordinary Coq function would not let us observe its use of comparisons.
-We make those uses observable by writing the algorithm as a monadically expressed function, parameterizing it on both the monad itself and on the comparison operation.
-Expressed this way, a deterministic Quicksort that simply selects the head of the input list as its pivot element, and uses two simple filter passes to partition the input list, looks as follows:
+Since we are working in Coq, we already have at our disposal a full functional
+programming language, in the form of Coq's pCIC \cite{coq'art}. However, just
+writing the algorithm as an ordinary Coq function would not let us observe its
+use of comparisons. 
+We can however see comparison counting as a \emph{side effect}. As is well
+known and standard practice in functional languages such as Haskell, side
+effects can be represented using \emph{monads}: a side-effecting function |f|
+from |A| to |B| is represented as a function |A -> M B| where |M| is a type
+constructor encapsulating the side effects. ``Identity'' and ``composition''
+for such functions are given by |ret| (for ``return'') of type |A -> M A| and
+|bind| (infix: |>>=|) of type |M A -> (A -> M B) -> M B| satisfying certain
+identities (the \emph{monad laws}). For a general introduction to monadic
+programming and monad laws, see \cite{wadler93monads}. Furthermore, we use
+Haskell's ``do-notation'',
+\begin{code}
+ Notation "x <- y ; z" := (bind y (fun x: _ => z)) 
+\end{code}
+and freely use standard monadic functions such as:
+\begin{code}
+ liftM: forall (M: Monad) (A B: Set), (A -> B) -> (M A -> M B)
+ filterM: forall (M: Monad) (A: Set), (A -> M bool) -> list A -> M (list A)
+\end{code}
+|Monad| is a dependent record containing the (coercible) carrier of type |Set
+-> Set|, along with the |bind| and |ret| operations, and proofs of the three monad laws. 
+% We make those uses observable by writing the algorithm as a monadically
+% expressed function, parameterizing it on both the monad itself and on the
+% comparison operation.
+
+We now express Quicksort in this style, parameterizing it on both the monad
+itself and on the comparison operation.
+A deterministic Quicksort that simply selects the head of the input list as its pivot element, and uses two simple filter passes to partition the input list, looks as follows:
 
@@
 \end{code}
-|Monad| is a dependent record containing the (coercible) carrier of type |Set -> Set|, along with |bind| (infix: |>>=|) and |ret| (for ``return'') operations, and the three monad laws.
-For a general introduction to monadic programming and monad laws, see \cite{wadler93monads}.
-Furthermore, we use Haskell's ``do-notation'', and freely use standard monadic functions such as |liftM| (of type |forall (M: Monad) (A B: Set), (A -> B) -> (M A -> M B)|) and |filterM| (of type |forall (M: Monad) (A: Set), (A -> M bool) -> list A -> M (list A)|).
 
@@
 By instantiating the above definitions with the right monad, we can transparently insert comparison-counting instrumentation into the algorithm, which will prove to be sufficient to let us reason about its complexity.
-But before we do so, let us note that if the above definitions are instead instantiated with the identity monad and an ordinary elementwise comparison on |T|, then the monadic scaffolding melts away, and the result is equivalent to an ordinary non-instrumented, non-monadic version, suitable for extraction and correctness proofs (which are included in the formalization for completeness).
+But before we do so, let us note that if the above definitions are instead instantiated with the identity monad and an ordinary elementwise comparison on |T|, then the monadic scaffolding melts away, and the result is equivalent to an ordinary non-instrumented, non-monadic version, suitable for extraction and correctness proofs (which are included in the formalisation for completeness).
 This means that while we will instantiate the definitions with less trivial monads to support our complexity proofs, we can take some comfort in knowing that the object of those proofs is, in a very concrete sense, the actual Quicksort algorithm (as one would write it in a functional programming language), rather than some idealized model thereof.
@@
 
-The Quicksort algorithm used by the average-case proof we formalize differs from the one presented in the last section in two ways.
-First, it selects a pivot element nondeterministically, thereby avoiding the pathological quadratic cases any deterministic pivot selection strategy inevitably suffers.
-Second, it uses a single three-way partition pass instead of two two-way filter passes, thereby avoiding the pathological quadratic case that two two-way filter passes suffer for an input list whose elements are all of the same value.
+The version of Quicksort used in the average-case proof in \cite{introtoalgos}
+differs from the one presented in the last section in two ways. This is also
+reflected in our formalisation.
 
-Combined, these two traits ensure that the $O(n \log n)$ average-case bound holds not just averaged over all possible input lists, but for each individual input list as well.
-This greatly simplifies things, because it removes the need to carefully track input distributions in order to show that nice inputs (for which the original deterministic version of the algorithm performs well) sufficiently outnumber bad inputs (for which the original deterministic version of the algorithm performs poorly).
-In particular, it means that once we prove that the bound holds for an arbitrary input, the global bound immediately follows.
-It also means that for a key lemma near the end of our proof, we can use straightforward induction over the algorithm's recursive call structure, without having to show that given appropriately distributed inputs, the partition step yields lists that are again appropriately distributed.
-The latter is a typical concern in more ambitious approaches for average-case complexity analysis \cite[for example]{modularcalculus} and analysis of probabilistic algorithms.
+First, the definition of |qs| is modified to use a single three-way partition
+pass, instead of two calls to |filter|, thus avoiding the pathological
+quadratic behaviour which can arise when the input list does not consist of
+distinct elements.
 
-While we sympathize with the view that changing the algorithm under investigation amounts to cheating, we will nevertheless simply follow the textbook proof (which represents common practice) in this regard.
-Hence, we now modify our |qs| definition to use a three-way partition pass, and to use nondeterministic pivot selection.
-Following that, we will compose a new monad (again using |MMT|, but this time transforming a nondeterminism monad) that the new definition can be instantiated with to capture its \emph{expected} comparison count.
+Second, and more significantly, we use \emph{nondeterministic} pivot
+selection, thus avoiding the pathological quadratic behaviour from which
+any deterministic pivot selection strategy inevitably suffers. While this
+means that we have proved our result for a subtly different presentation of
+Quicksort, this nevertheless follows the textbook treatment, in line with
+common practice.
 
-First, we define |partition|.
-Instead of a two-way comparison operation producing a monadic |bool|, it takes a three-way comparison operation producing a monadic |comparison|, which is an enumeration with values |Lt|, |Eq|, and |Gt|.
-We represent the resulting partitioning by a function of type |comparison -> list T| rather than a record or tuple type containing three lists, because in the actual formalization, this saves us from having to constantly map |comparison| values to corresponding record field accessors or tuple projections.
+These two modifications together greatly simplify the formalisation, because
+they remove the need to carefully track input distributions in order to show
+that nice inputs (for which the original deterministic version of the
+algorithm performs well) sufficiently outnumber bad inputs (for which the
+original version performs poorly). They further ensure that the $O(n \log n)$
+average-case bound holds not just averaged over all possible input lists, but
+for each individual input list as well. In particular, it means that once we
+prove that the bound holds for an arbitrary input, the global bound
+immediately follows.
+
+This also means that for a key lemma near the end of our proof, we can use
+straightforward induction over the algorithm's recursive call structure,
+without having to show that given appropriately distributed inputs, the
+partition step yields lists that are again appropriately distributed. Such
+issues are a major technical concern in more ambitious approaches to
+average-case complexity analysis \cite[for example]{modularcalculus} and to
+the analysis of probabilistic algorithms.
+
+The second modification is based on a new monad (again defined using |MMT|,
+but this time transforming a nondeterminism monad) with which the new
+definition can be instantiated, capturing the \emph{expected} comparison
+count.
+
+The first modification is relatively straightforward. Instead of calling
+|filterM|, which uses a two-way comparison operation producing a monadic |bool|, we
+define a function |partition|.
+It takes a three-way comparison operation producing a monadic |comparison|, which is an enumeration with values |Lt|, |Eq|, and |Gt|.
+We represent the resulting partitioning by a function of type |comparison -> list T| rather than a record or tuple type containing three lists, because in the actual formalisation, this saves us from having to constantly map |comparison| values to corresponding record field accessors or tuple projections.
 This is only a matter of minor convenience; a record or tuple could have been used instead without problems.
@@
 Note that the deterministic Quicksort definition in section \ref{embed} could have been implemented with a |partition| pass as well, which might well have made the worst-case proof even simpler.
-We chose not to do this, in order to emphasize that the properties the average-case proof demands of the algorithm rule out the naive but familiar implementation using |filter| passes.
+We chose not to do this, in order to emphasize that the properties the average-case proof demands of the algorithm rule out the na{\"\i}ve but familiar implementation using |filter| passes.
 
@@
   |expec f (t >>= (ret fcmp g))| &= |expec (f fcmp g) t| \nonumber \\
-    &= |expec f (ne_tree.map g t)| \label{expec_bind_lem}
+  |expec (f fcmp g) t| &= |expec f (ne_tree.map g t)| \label{expec_bind_lem}
 \end{align}
@@
 The first equality expresses that the expected number of comparisons counted by |NDP| is equal to the expected length of the list of comparisons recorded by |U|.
-In the formalization, this is a separate lemma proved by induction on |qs|.
+In the formalisation, this is a separate lemma proved by induction on |qs|.
 The second equality merely unfolds the definition of |monoid_expec|.
 
-After rewriting with identity \ref{expec_bind_lem} from section \ref{nondetexpec}, the goal becomes
+After rewriting with identity \ref{expec_bind_lem} in section \ref{nondetexpec} on page \pageref{expec_bind_lem}, the goal becomes
 \begin{code}
@@
 \end{code}
-Since |cmp_U| is deterministic, |partition_U| is as well.
-Furthermore, since we know exactly what monadic effects |partition_U| has, we can split those effects off and revert to simple uneffectful |filter| passes.
-Finally, we rewrite using the following |monoid_expec| identity:
+Since |cmp_U| is deterministic, |partition_U| is as well. Furthermore, since
+we know exactly what monadic effects |partition_U| has, we can split those
+effects off and revert to simple effect-free |filter| passes. Finally, we rewrite using the following |monoid_expec| identity:
 \begin{code}
@@
 In the interest of brevity, we have omitted lots of detail and various lemmas in the description of the proof.
-Still, the parts shown are reasonably faithful to the actual formalization, with two notable exceptions.
+Still, the parts shown are reasonably faithful to the actual formalisation, with two notable exceptions.
 
-First, we have pretended to have used ordinary natural numbers as indices into ordinary lists, completely ignoring issues of index validity that could not be ignored in the actual formalization.
+First, we have pretended to have used ordinary natural numbers as indices into ordinary lists, completely ignoring issues of index validity that could not be ignored in the actual formalisation.
 There, we use vectors (lists whose size is part of their type) and bounded natural numbers in many places instead.
@@
 Since the recursive calls are nested in lambda abstractions passed to the |bind| operation of an unspecified monad, the relation between their arguments and the function's parameters is not locally known, resulting in unprovable proof obligations.
-To make these provable, we $\Sigma$-decorated the types of |filter| and |partition| in the actual formalization with modest length guarantees.
+To make these provable, we $\Sigma$-decorated the types of |filter| and |partition| in the actual formalisation with modest length guarantees.
 % \item Using the induction principles corresponding to the definitions produced by |Program| for a function already defined with two layers of monadicity quickly became unwieldy.
@@
 
-The formalized development successfully adopted from the original proof the idea of using a nondeterministic version of the algorithm to make the $O(n \log n)$ bound hold for any input list, the idea of taking an order-indexed perspective to reduce the problem to a sum-of-expected-comparison-counts, and the use of the standard bound for harmonic series for the strictly numerical part.
+The formalised development successfully adopted from the original proof the idea of using a nondeterministic version of the algorithm to make the $O(n \log n)$ bound hold for any input list, the idea of taking an order-indexed perspective to reduce the problem to a sum-of-expected-comparison-counts, and the use of the standard bound for harmonic series for the strictly numerical part.
 However, for the actual reduction and the derivation of the formula for the expected comparison count, the intuitive arguments essentially had to be reworked from scratch, building on the monadic representation of the algorithm and the various comparison counting/nondeterminism monads.

### 2009-02-19 paper/compile.sh
paper/compile.sh: Save processed.tex.
Ignore-this: 37d7815babc3e10a8b22479e890327ca

darcs-hash:20090219080340-bab43-78f4ae5f8e004617e1f59f1c3275b68edf584917.gz

@@ -1,12 +1,13 @@
 #!/bin/sh
 
 lhs2TeX < paper.tex > processed.tex
+cp processed.tex tmp.tex
 
 set -e
-pdflatex processed.tex
-bibtex processed
-pdflatex processed.tex
-bibtex processed
-pdflatex processed.tex
-cat processed.pdf > paper.pdf
-rm processed.*
+pdflatex tmp.tex
+bibtex tmp
+pdflatex tmp.tex
+bibtex tmp
+pdflatex tmp.tex
+cat tmp.pdf > paper.pdf
+rm tmp.*

### 2009-02-19 paper/paper.tex
paper: Fix inverted inequality.
Ignore-this: d6a98221868499b0a1f6ff1a00f269fa

darcs-hash:20090219084636-bab43-62012aace058b21a63e5711102bae7872c12bde1.gz
james-final-tweaks
darcs-hash:20090219132958-d98fc-d81cd9874759ac4b5fafca435eef76905f7f3dad.gz

@@
 The key to giving a proper formal treatment of both of these aspects lies in the use of an appropriate representation of the algorithm, capable of capturing its computational behavior---specifically, its use of comparisons---in a way suitable for subsequent formal reasoning.
-The representation we use takes the form of a shallow monadic embedding, in which the algorithm is written as a monadically expressed functional program.
+The appproach we take is to consider such operation-counting as a \emph{side effect}, and to use the general framework of \emph{monads} for representing side-effecting computation in pure functional languages.
+The representation we use takes the form of a shallow monadic embedding, in which the algorithm, here Quicksort, is written as a monadically expressed functional program.
 This definition is then instantiated with refinements of operation-counting monads to make the comparison count observable.
@@
 \end{code}
-|Monad| is a dependent record containing the (coercible) carrier of type |Set
+Here, the Coq type |Monad| is a dependent record containing the (coercible) carrier of type |Set
 -> Set|, along with the |bind| and |ret| operations, and proofs of the three monad laws. 
@@
 We use Coq's |Program Fixpoint| facility \cite{subsetcoercions} to cope with Quicksort's non-structural recursion, specifying list length as an input measure function that is separately shown to strongly decrease for each recursive call.
-For |qs|, these proof obligations are trivial enough for Coq to prove mostly by itself.
+For this definition of |qs|, these proof obligations are trivial enough for Coq to prove mostly by itself.
 
 For recursive functions defined this way, Coq does not automatically define corresponding induction principles matching the recursive call structure.
-Hence, for this |qs| definition as well as the one we will introduce in section \ref{nondetexpec}, we have defined these induction principles manually.
+Hence, for this |qs| definition as well as the one we will introduce in section \ref{nondetexpec}, we had to define these induction principles manually.
 To make their use as convenient as possible, we further customized and specialized them to take advantage of specific monad properties.
-We will ignore these issues in this paper, and will henceforth simply say: ``by induction on |qs|, ...''.
+We will omit further discussion of these issues in this paper, and will henceforth simply say: ``by induction on |qs|, ...''.
 
@@
 
-We now expand the technique in preparation for the average-case complexity proof.
+We now extend the technique to prepare for the average-case complexity proof.
 
@@
 they remove the need to carefully track input distributions in order to show
-that nice inputs (for which the original deterministic version of the
-algorithm performs well) sufficiently outnumber bad inputs (for which the
+that `good' inputs (for which the original deterministic version of the
+algorithm performs well) sufficiently outnumber `bad' inputs (for which the
 original version performs poorly). They further ensure that the $O(n \log n)$
@@
 
-Note that the deterministic Quicksort definition in section \ref{embed} could have been implemented with a |partition| pass as well, which might well have made the worst-case proof even simpler.
+Note that the deterministic Quicksort definition in section \ref{embed} could also have been implemented with a |partition| pass instead, which might well have made the worst-case proof even simpler.
 We chose not to do this, in order to emphasize that the properties the average-case proof demands of the algorithm rule out the na{\"\i}ve but familiar implementation using |filter| passes.
@@
 Nondeterminism can now be emulated by instantiating these definitions with a suitable monad and |pick| operation.
-A deterministic, non-instrumented version can still be obtained, simply by using the identity monad and a deterministic |pick| operation, like |head|.
+A deterministic, non-instrumented version can still be obtained, simply by using the identity monad and any deterministic |pick| operation, such as |head| or 'median-of-three'.
 
@@
 produces |[0, 1, 2]| as its list of possible outcomes.
-Unfortunately, the information that 0 is a more likely outcome than 1 or 2, has been lost.
+Unfortunately, the information that 0 is a more likely outcome than 1 or 2 has been lost.
 Such relative probabilities are critical to the notion of an expected value: the expected value of the program above is |avg [0, avg [1, 2]] = {-"\frac{3}{4}"-} /= 1 = avg [0, 1, 2]|.
@@
 
-The right inequality is a strictly numerical affair, requiring a bit of analysis involving the harmonic series.
-As stated before, this part of the proof could be fairly directly transcribed from the paper proof, with few complications and additions, and so we will not discuss it.
+The right inequality is a strictly numerical affair, requiring a bit of analysis involving the harmonic series. As stated before, this part of the proof 
+    was 
+%%% could be 
+fairly directly transcribed from the paper proof, with few complications and additions, and so we will not discuss it.
 
@@
 We end up with two subgoals, the first of which is \begin{code}forall (i, j) `NotIn` IJ, expec (count (i, j)) (ne_tree.map fst (qs_U li)) = 0.\end{code} Rewriting this using identity \ref{expec_bind_lem} from section \ref{nondetexpec} in reverse, then rewriting the |expec| as a |monoid_expec|, and then generalizing the premise, results in
+%%% this needs a label!!! (\label{monoid_expec_zero_lem})
 \begin{code}
-  forall i j li, (i `NotIn` li \/ j `NotIn` li) -> monoid_expec (count (i, j)) (U.qs li) = 0,
+  forall i j li, (i `NotIn` li \/ j `NotIn` li) -> monoid_expec (count (i, j)) (U.qs li) = 0 
 \end{code}
@@
 
-The second subgoal, expressed with |monoid_expec|, reads
+The second subgoal, expressed with |monoid_expec|, becomes
+%%% this needs a label!!! (\label{monoid_expec_IJ_lem})
 \begin{code}
@@
 \end{code}
-This corresponds exactly to the property described at the beginning of this section. We prove it in the next section.
+which corresponds exactly to the property described at the beginning of this section. We prove it in the next section.
 
@@
 
-Again, the proof is by induction on |qs|.
+Again, the proof %%% of (\ref{monoid_expec_IJ_lem}) 
+is by induction on |qs|.
 But to get a better induction hypothesis, we drop the |(i, j) `In` IJ| premise (because as was shown in the last section, the statement is also true if |(i, j) `NotIn` IJ|), and add a premise saying |li| is a permutation of a contiguous sequence of indices.
@@
 Since the former precede the latter in |sort l|, it must be the case that these indices are all less than |pivot|.
-And since |pivot < i|, it follows that the first |qs_U| term will produce no |(i, j)| comparisons (using the property shown at the end of the previous section).
-Hence, the first |monoid_expec| term vanishes as well, leaving
+And since |pivot < i|, it follows that the first |qs_U| term will produce no |(i, j)| comparisons (using the property
+%%%(\ref{monoid_expec_IJ_lem}) 
+shown at the end of the previous section). Hence, the first |monoid_expec| term vanishes as well, leaving
 \begin{code}
@@
 \end{itemize}
-This concludes the case where |i < pivot|. The case where |j < pivot| is symmetric. The other three cases use similar arguments.
+This concludes the case where |pivot < i|. The case where |j < pivot| is symmetric. The other three cases use similar arguments.
 
@@
 We have shown that it is straightforward to give a fully formal treatment in type theory of a classical result in complexity theory.
-This clearly shows the utility and applicability of the general monadic framework for worst- and average-case complexity that we have developed.
+This clearly shows the utility and applicability of the general monadic
+%%%framework for 
+approach to worst- and average-case complexity that we have developed.
 

### 2009-02-19 paper/paper-unlabelled.tex
labelling equations
darcs-hash:20090219142844-d98fc-bae29ff836240fda827e321c94d273c24149f733.gz

--- 2009-02-19 paper/paper.tex
@@
 Nondeterminism can now be emulated by instantiating these definitions with a suitable monad and |pick| operation.
-A deterministic, non-instrumented version can still be obtained, simply by using the identity monad and any deterministic |pick| operation, such as |head| or 'median-of-three'.
+A deterministic, non-instrumented version can still be obtained, simply by using the identity monad and any deterministic |pick| operation, such as |head| or 'median-of-three' (not considered here).
 

### 2009-02-19 paper/paper.tex
final 16 page tweaks plus bib tidy plus labels
darcs-hash:20090219160748-d98fc-32e0c5d1ee487b0bfb22dbcaa8a3e8036b6e8e65.gz

@@
 Indeed, it does not even formally define the algorithm in the first place, relying instead on assertions which are taken to be intuitively true.
-While this practice is common and perfectly reasonable for paper proofs intended for human consumption, it is a luxury we do not afford ourselves.
+While this practice is common and perfectly reasonable for paper proofs intended for human consumption, it is a luxury we can not afford ourselves.
 
 Second, the original proof (implicitly) assumes that the input list does not contain any duplicate elements, which significantly simplifies its derivation of the formula for the expected comparison count for pairs of individual input list elements.
-We do not make this assumption.
+We take care to avoid appeals to such an assumption.
   % todo: perhaps bring this up later.
 
-The key to giving a proper formal treatment of both of these aspects lies in the use of an appropriate representation of the algorithm, capable of capturing its computational behavior---specifically, its use of comparisons---in a way suitable for subsequent formal reasoning.
+The key to giving a proper formal treatment of both these aspects lies in using an appropriate representation of the algorithm, capable of capturing its computational behavior---specifically, its use of comparisons---in a way suitable for subsequent formal reasoning.
 The appproach we take is to consider such operation-counting as a \emph{side effect}, and to use the general framework of \emph{monads} for representing side-effecting computation in pure functional languages.
-The representation we use takes the form of a shallow monadic embedding, in which the algorithm, here Quicksort, is written as a monadically expressed functional program.
+Accordingly we use a shallow embedding, in which the algorithm, here Quicksort, is written as a monadically expressed functional program in Coq.
 This definition is then instantiated with refinements of operation-counting monads to make the comparison count observable.
@@
 The next two sections detail the actual formalised proof.
-Section \ref{reduction} corresponds to the first part in the original proof described above, and shows how the main theorem can be split up into a lemma (stated in terms of another specialized monad) giving a formula for the expected comparison count for individual pairs of input element, and a strictly numerical part.
-Since we were able to fairly directly transcribe the latter part from the paper proof, using the existing real number theory in the Coq standard library with few complications and additions, we omit a discussion of it in this paper and refer the interested reader to the paper proof.
+Section \ref{reduction} corresponds to the first part in the original proof described above, showing how the main theorem can be split into a lemma (stated in terms of another specialized monad) giving a formula for the expected comparison count for individual pairs of input elements, and a strictly numerical part.
+Since we were able to fairly directly transcribe the latter from the paper proof, using the existing real number theory in the Coq standard library with few complications and additions, we omit discussion of it here and refer the interested reader to the paper proof.
 
@@
 
-The Coq source files containing the entire formalisation can be downloaded from \url{http://www.eelis.net/research/quicksort/}. The Coq version used is 8.2 beta 2.
+The Coq source files containing the entire formalisation can be downloaded from \url{http://www.eelis.net/research/quicksort/}. We used Coq version 8.2 beta 2.
 
@@
 A completely different approach to type-theoretic analysis of computational complexity is to devise a special-purpose type theory in which the types of terms include some form of complexity guarantees.
-This is the approach taken in \cite{constable}, for example.
+Such an approach is taken in \cite{constable}, for example.
 
@@
 identities (the \emph{monad laws}). For a general introduction to monadic
-programming and monad laws, see \cite{wadler93monads}. Furthermore, we use
-Haskell's ``do-notation'',
+programming and monad laws, see \cite{wadler93monads}.
+
+\noindent Furthermore, we use
+Haskell's ``do-notation'', declared in Coq as follows 
 \begin{code}
@@
 \end{code}
-
+%%%
 %format `infixbind_SP` = >>="_{SP}"
-
+%%%
 Defining |cost| and |result| as the first and second projection, respectively, we trivially have identities such as |cost (ret_SP x) = 0|, |cost (le_SP x y) = 1|, and |cost (x `infixbind_SP` f) = cost x + cost (f (result x))|.
@@
 
-We now extend the technique to prepare for the average-case complexity proof.
+We now extend the technique to prepare for the average-case proof.
 
@@
 
-The version of Quicksort used in the average-case proof in \cite{introtoalgos}
+The version of Quicksort used in the average-case complexity proof in \cite{introtoalgos}
 differs from the one presented in the last section in two ways. This is also
@@
 \end{code}
-|qs_U| operates directly on lists of indices into |sort l|.
-Comparison of indices is defined by comparison of the |T| values they denote in |sort l|.
-Furthermore, rather than producing a grand total comparison count the way |NDP| does, |U| records every pair of indices compared, by using |MMT| with |Monoid_U|, the free monoid over |nat prod nat| pairs, instead of the additive monoid on |nat| we used up until now.
+The function |qs_U| operates directly on lists of indices into |sort l|. Comparison of indices is defined by comparison of the %%%|T| 
+values they denote in |sort l|.
+Furthermore, rather than producing a grand total comparison count the way |NDP| does, |U| records every pair of indices compared, by using |MMT| with |Monoid_U|, the free monoid over |nat prod nat| pairs, instead of the additive monoid on |nat| we used until now.
 
@@
 %%% this needs a label!!! (\label{monoid_expec_zero_lem})
-\begin{code}
-  forall i j li, (i `NotIn` li \/ j `NotIn` li) -> monoid_expec (count (i, j)) (U.qs li) = 0 
-\end{code}
-which can be shown by induction on |qs|, although we will not do so in this paper. We \emph{will} use this property again in the next section.
+\begin{align}
+  |forall i j li, (i `NotIn` li \/ j `NotIn` li) -> monoid_expec (count (i, j)) (U.qs li) = 0| \label{monoid_expec_zero_lem}
+\end{align}
+which can be shown by induction on |qs|, although we will not do so in this paper. 
+We \emph{will} use this property again in the next section.
 
@@
 %%% this needs a label!!! (\label{monoid_expec_IJ_lem})
-\begin{code}
-  forall (i, j) `In` IJ, monoid_expec (count (i, j)) (qs_U li) <= ecc i j. 
-\end{code}
+\begin{align}
+  |forall (i, j) `In` IJ, monoid_expec (count (i, j)) (qs_U li) <= ecc i j| \label{monoid_expec_IJ_lem}
+\end{align}
 which corresponds exactly to the property described at the beginning of this section. We prove it in the next section.
@@
 
-Again, the proof %%% of (\ref{monoid_expec_IJ_lem}) 
+Again, the proof of (\ref{monoid_expec_IJ_lem}) 
 is by induction on |qs|.
@@
 \item the case where the |pivot| is equal to |i| occurs once, and in this case no more than a single |(i, j)| comparison is expected;
-\item in the case where |pivot| lies between |i| and |j|, the number of expected |(i, j)| comparisons is 0, and consequently it does not matter how many times this case occurs;
+\item in the case where |pivot| lies between |i| and |j|, the number of expected |(i, j)| comparisons is 0, and hence it does not matter how often this case occurs;
 \item the case where the |pivot| is equal to |j| occurs once, and in this case no more than a single |(i, j)| comparison is expected;
@@
 \end{code}
-
-Five subgoals remain after applying the above lemma---one for each of the listed cases. The first one reads
+%%%
+Five subgoals remain after applying 
+%%% the above 
+    this 
+lemma---one for each 
+%%% of the listed cases. 
+    listed case. 
+The first one reads
 \begin{code}
@@
 \end{code}
-All indices in the first filtered list denote elements that are less than the element denoted by the pivot.
+All indices in the first filtered list denote elements 
+%%% that are 
+less than the element denoted by the pivot.
 Since the former precede the latter in |sort l|, it must be the case that these indices are all less than |pivot|.
-And since |pivot < i|, it follows that the first |qs_U| term will produce no |(i, j)| comparisons (using the property
-%%%(\ref{monoid_expec_IJ_lem}) 
-shown at the end of the previous section). Hence, the first |monoid_expec| term vanishes as well, leaving
+And since |pivot < i|, it follows that the first |qs_U| term will produce no |(i, j)| comparisons (using property
+   (\ref{monoid_expec_zero_lem})  
+shown at the end of section~\ref{reduction}). Hence, the first |monoid_expec| term vanishes%%% 
+%%% as well
+, leaving
 \begin{code}
@@
 This concludes the case where |pivot < i|. The case where |j < pivot| is symmetric. The other three cases use similar arguments.
-
+%%%
 The proof is now complete.
@@
 Being parameterized on the monad used, it allows a single definition to be instantiated either with basic monads (like the identity monad or bare nondeterminism monads) to get a non-instrumented version suitable for extraction and correctness proofs, or with |MMT|-transformed monads to support complexity proofs.
-Furthermore, since the shallowness of the embedding lets us re-use all standard Coq data types and facilities, including the powerful |Program Fixpoint| command, the actual algorithm definition itself is reasonably clean.
+Furthermore, since
+%%% the shallowness of the embedding 
+    this approach lets us re-use all standard Coq data types and facilities, including the powerful |Program Fixpoint| command, the actual algorithm definition itself is reasonably clean.
 
@@
 This clearly shows the utility and applicability of the general monadic
-%%%framework for 
-approach to worst- and average-case complexity that we have developed.
+%%% framework for 
+    approach 
+%%% to worst- and average-case complexity that 
+we have developed.
 

### 2009-02-19 paper/paper.tex
ispell -b for british spelling
darcs-hash:20090219162103-d98fc-037b1406047bacd85202f9131518a43a1c1c0619.gz

@@
 
-The key to giving a proper formal treatment of both these aspects lies in using an appropriate representation of the algorithm, capable of capturing its computational behavior---specifically, its use of comparisons---in a way suitable for subsequent formal reasoning.
-The appproach we take is to consider such operation-counting as a \emph{side effect}, and to use the general framework of \emph{monads} for representing side-effecting computation in pure functional languages.
+The key to giving a proper formal treatment of both these aspects lies in using an appropriate representation of the algorithm, capable of capturing its computational behaviour---specifically, its use of comparisons---in a way suitable for subsequent formal reasoning.
+The approach we take is to consider such operation-counting as a \emph{side effect}, and to use the general framework of \emph{monads} for representing side-effecting computation in pure functional languages.
 Accordingly we use a shallow embedding, in which the algorithm, here Quicksort, is written as a monadically expressed functional program in Coq.
@@
 Section \ref{finishing} finishes the proof by proving the lemma about the expected comparison count for individual input list elements.
-Since this is the part where the original proof omits the most detail, and makes the assumption regarding duplicate elements, and where we really have to reason in detail about the behavior of the algorithm, it is by far the most involved part of the formalisation.
+Since this is the part where the original proof omits the most detail, and makes the assumption regarding duplicate elements, and where we really have to reason in detail about the behaviour of the algorithm, it is by far the most involved part of the formalisation.
 
@@
 Note that the deterministic Quicksort definition in section \ref{embed} could also have been implemented with a |partition| pass instead, which might well have made the worst-case proof even simpler.
-We chose not to do this, in order to emphasize that the properties the average-case proof demands of the algorithm rule out the na{\"\i}ve but familiar implementation using |filter| passes.
+We chose not to do this, in order to emphasise that the properties the average-case proof demands of the algorithm rule out the na{\"\i}ve but familiar implementation using |filter| passes.
 
@@
 \end{document}
+
+% LocalWords:  Nijmegen GL Quicksort Coq todo monadically Hurd Audebaud Paulin
+% LocalWords:  Mohring na ve Coq's pCIC ret Haskell's liftM forall filterM bool
+% LocalWords:  parameterizing le gt negb Fixpoint qs Quicksort's customized MMT
+% LocalWords:  specialized elementwise idealized monoid fcmp ab snd mult fst SP
+% LocalWords:  nat NatAddMonoid IdMonad Eval infixbind sqrd flt Eq cmp const ne
+% LocalWords:  upp avg ary expec NDP MonoidMonadTrans NeTreeMonad bigO INR ceil
+% LocalWords:  IJ ecc li fr NotIn generalizing ca cb foo parameterized

### 2009-02-19 paper/paper.tex
another bad "'"!
darcs-hash:20090219174744-d98fc-dbefa4cc57dc2f55b86e1d2f3bd933abbda7fc99.gz

@@
 Nondeterminism can now be emulated by instantiating these definitions with a suitable monad and |pick| operation.
-A deterministic, non-instrumented version can still be obtained, simply by using the identity monad and any deterministic |pick| operation, such as |head| or 'median-of-three' (not considered here).
+A deterministic, non-instrumented version can still be obtained, simply by using the identity monad and any deterministic |pick| operation, such as |head| or `median-of-three' (not considered here).
 

### 2009-02-19 paper/paper.tex
boilerplate tweaks
darcs-hash:20090219181245-d98fc-70b3ab13787a804da62441ff71e27b1a7248e2da.gz

@@
 
-\author{\eelis\thanks{Research carried out by the first author as part of the Master's programme in ``Foundations''} and \james \\
+\author{\eelis\thanks{Research carried out as part of the Radboud Master's programme in ``Foundations''} and \james \\
 eelis@@eelis.net, james.mckinna@@cs.ru.nl}
 
-\institute{Department of Computer Science, University of Nijmegen \\
-P.O. Box 9010, 6500 GL Nijmegen, The Netherlands}
+\institute{
+Institute for Computing and Information Sciences \\
+Radboud University Nijmegen \\
+Heijendaalseweg 135, 6525 AJ Nijmegen, The Netherlands}
 

### 2009-02-23 paper/paper.tex
bib fiddling for springer
darcs-hash:20090220082649-d98fc-dae1df0194319b97160e3bab21e92b4404d48fb5.gz
paper: Specify Coq version 8.2.
Ignore-this: 3643ada754cd6280cac9daa13079a9b1

darcs-hash:20090220091248-bab43-05a80293b809e60f9a8ec0e1e238a3c9e5e9dd06.gz
paper: Correct Wadler reference, erase dangling Jones reference.
Ignore-this: 4fb42d2af3bdd6d876a741994ac111cb

darcs-hash:20090220135553-bab43-3546114a04ea4093f08e875af56930da56765195.gz
paper: Use CIC instead of pCIC. (There is no justice.)
Ignore-this: b04278ad5ea29cbfe0fb3e3f5972da88

darcs-hash:20090223154803-bab43-ffaf197fd1620f5946909

@@
 
-The Coq source files containing the entire formalisation can be downloaded from \url{http://www.eelis.net/research/quicksort/}. We used Coq version 8.2 beta 2.
+The Coq source files containing the entire formalisation can be downloaded from \url{http://www.eelis.net/research/quicksort/}. We used Coq version 8.2.
 
@@
 Since we are working in Coq, we already have at our disposal a full functional
-programming language, in the form of Coq's pCIC \cite{coq'art}. However, just
+programming language, in the form of Coq's CIC \cite{coq'art}. However, just
 writing the algorithm as an ordinary Coq function would not let us observe its
@@
 identities (the \emph{monad laws}). For a general introduction to monadic
-programming and monad laws, see \cite{wadler93monads}.
+programming and monad laws, see \cite{wadler95monads}.
 
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{splncs}
 \addcontentsline{toc}{chapter}{Bibliography}

### 2009-02-24 scripts/make_tarball.sh
Add scripts/make_tarball.sh.
Ignore-this: d5a63c357dda557a1cecd3fda54d169d

darcs-hash:20090224085324-bab43-652159b01c17eddaf8f1bccaa3e62416ec5ccdff.gz

@@ -1,0 +1,5 @@
+#!/bin/bash
+set -ev
+cd ..
+rm -f qs-avg.tar.gz
+tar -czf qs-avg.tar.gz qs-avg/src/{*.v,README.html,SConstruct} qs-avg/paper/{abstract.ltx,compile.sh,llncs.cls,macros.sty,paper.bib,paper.pdf,paper.tex,processed.tex,splncs.bst}

### 2009-02-24 paper/paper.tex
Yet more final tweaks.
Ignore-this: 200941a53a064f208d2b0bbcc35b4cd0

darcs-hash:20090224143438-bab43-74608dd14c2310bd1b6a55f84e6d46c6113f2607.gz

@@
 
-For reasons that will become clear in later sections, we compose the monad we will instantiate the above definitions with using a monad transformer \cite{monadtrans} |MMT| (for ``monoid monad transformer''), which piggybacks a monoid onto an existing monad by pairing.
+For reasons that will become clear in later sections, we construct the monad with which we will instantiate the above definitions using a monad transformer \cite{monadtrans} |MMT| (for ``monoid monad transformer''), which piggybacks a monoid onto an existing monad by pairing.
 
@@
 
-We now use |MMT| to piggyback the additive monoid on |nat| onto the identity monad, and lift elementwise comparison into the resulting monad, which we call |SP| (for ``simply-profiled'').
+We now use |MMT| to piggyback the additive monoid structure on |nat| onto the identity monad, and lift elementwise comparison into the resulting monad, which we call |SP| (for ``simply-profiled'').
 \begin{code}
@@
 
-To denote the expected value of a \emph{measure} of the output of a program, we define
+To denote the expected value of a discrete measure |f| of the output of a program, we define
 
@@
 
-The last thing needed before the main theorem can be stated, is the notion of big-O complexity. We use a standard textbook definition, except that we incorporate a measure function parameter |m|:
+The last thing needed before the main theorem can be stated, is the notion of big-O complexity.
+We use the standard textbook definition, except that we make explicit how we measure inputs to |f|, namely with respect to a measure function |m|:
 
@@
 
-  Notation {-"``"-}over m,{-"\ "-} f =O( g ){-"\text{''''}"-} := bigO m f g.
+  Notation {-"``"-}wrt m,{-"\ "-} f =O( g ){-"\text{''''}"-} := bigO m f g.
 \end{code}
@@
 \begin{code}
-  Theorem qs_avg: over length, expec cost fcmp qs_NDP =O(fun n => INR (n * log2ceil n)).
+  Theorem qs_avg: wrt length, expec cost fcmp qs_NDP =O(fun n => INR (n * log2ceil n)).
 \end{code}
@@
    (\ref{monoid_expec_zero_lem})  
-shown at the end of section~\ref{reduction}). Hence, the first |monoid_expec| term vanishes%%% 
+at the end of section~\ref{reduction} on page \pageref{monoid_expec_zero_lem}). Hence, the first |monoid_expec| term vanishes%%%
 %%% as well

### 2009-02-24 paper/paper.tex
Final tweaks (honestly).
Ignore-this: 8f6fbcf924041e9f7123a040b692ccf

darcs-hash:20090224144219-bab43-41dddddf608c66f18651c9380b6288affda59c6e.gz

@@
 
-The right inequality is a strictly numerical affair, requiring a bit of analysis involving the harmonic series. As stated before, this part of the proof 
+The right-hand inequality is a strictly numerical affair, requiring a bit of analysis involving the harmonic series. As stated before, this part of the proof 
     was 
@@
 The left inequality is the challenging one.
-To bring it closer to the index summation, we first write |l| on the left side as |map (nth (sort l)) li|, where |sort| may be any sorting function (including |qs| itself), and where |li| is a permutation of |[0...n-1]| such that |map (nth (sort l)) li = l| (such an |li| can easily be proven to exist).
+To bring it closer to the index summation, we first write |l| on the left-hand side as |map (nth (sort l)) li|, where |sort| may be any sorting function (including |qs| itself), and where |li| is a permutation of |[0...n-1]| such that |map (nth (sort l)) li = l| (such an |li| can easily be proven to exist).
 
@@
 \end{code}
-In the base case, |li| is |nil|, and the left side of the inequality reduces to 0. In the recursive case, |qs| unfolds:
+In the base case, |li| is |nil|, and the left-hand side of the inequality reduces to 0. In the recursive case, |qs| unfolds:
 \begin{code}
@@
 Each case occurs a certain number of times, and has an associated expected number of |(i, j)| comparisons (coming either from the |map_fst| term representing the |partition| pass, or from the two recursive |qs_U| calls).
-To represent this split, we first rewrite the right side of the inequality to\[\frac{\Varid{ecc}\ i\ j * (i - b) + 1 + 0 + 1 + \Varid{ecc}\ i\ j * (b + n - j)}{n}\ .\]
+To represent this split, we first rewrite the right-hand side of the inequality to\[\frac{\Varid{ecc}\ i\ j * (i - b) + 1 + 0 + 1 + \Varid{ecc}\ i\ j * (b + n - j)}{n}\ .\]
 This form reflects the facts that
@@
 \end{itemize}
-With the right side of the inequality in this form, we unfold the |avg| application on the left into |sum (...) / n|, and then cancel the division by |n| on both sides.
+With the right-hand side of the inequality in this form, we unfold the |avg| application on the left into |sum (...) / n|, and then cancel the division by |n| on both sides.
 Next, to actually realize the split, we apply a specialized lemma stating that

### 2009-02-24 scripts/make_tarball.sh
Add paper/README.txt.
Ignore-this: 2a8a02bd26f03c43b5b69b16be8535bf

darcs-hash:20090224144509-bab43-06a23d1a91d745f7c5e656603ca91cb0ccc0aeeb.gz

@@ -2,4 +2,4 @@
 set -ev
 cd ..
 rm -f qs-avg.tar.gz
-tar -czf qs-avg.tar.gz qs-avg/src/{*.v,README.html,SConstruct} qs-avg/paper/{abstract.ltx,compile.sh,llncs.cls,macros.sty,paper.bib,paper.pdf,paper.tex,processed.tex,splncs.bst}
+tar -czf qs-avg.tar.gz qs-avg/src/{*.v,README.html,SConstruct} qs-avg/paper/{README.txt,abstract.ltx,compile.sh,llncs.cls,macros.sty,paper.bib,paper.pdf,paper.tex,processed.tex,splncs.bst}

### 2010-04-10 papers/TYPES08/paper-labelled.tex
Minor repo reorganization. Remove website from repo.
Ignore-this: 6315eb60d8bd06ea4e105667cce0f8c5

darcs-hash:20100410124808-bab43-2c7f4ade6f1fd4ae62cedf7c97e058f96fabdcda.gz

--- 2009-02-19 paper/paper-unlabelled.tex
@@
 %%% this needs a label!!! (\label{monoid_expec_zero_lem})
-\begin{code}
-  forall i j li, (i `NotIn` li \/ j `NotIn` li) -> monoid_expec (count (i, j)) (U.qs li) = 0 
-\end{code}
+\begin{align}
+  |forall i j li, (i `NotIn` li \/ j `NotIn` li) -> monoid_expec (count (i, j)) (U.qs li) = 0| \label{monoid_expec_zero_lem}
+\end{align}
 which can be shown by induction on |qs|, although we will not do so in this paper. We \emph{will} use this property again in the next section.
@@
 Since the former precede the latter in |sort l|, it must be the case that these indices are all less than |pivot|.
-And since |pivot < i|, it follows that the first |qs_U| term will produce no |(i, j)| comparisons (using the property
-%%%(\ref{monoid_expec_IJ_lem}) 
+And since |pivot < i|, it follows that the first |qs_U| term will produce no |(i, j)| comparisons (using property
+   (\ref{monoid_expec_zero_lem})  
 shown at the end of the previous section). Hence, the first |monoid_expec| term vanishes as well, leaving

### 2010-04-10 scripts/make_tarball.sh

@@ -1,5 +1,0 @@
-#!/bin/bash
-set -ev
-cd ..
-rm -f qs-avg.tar.gz
-tar -czf qs-avg.tar.gz qs-avg/src/{*.v,README.html,SConstruct} qs-avg/paper/{README.txt,abstract.ltx,compile.sh,llncs.cls,macros.sty,paper.bib,paper.pdf,paper.tex,processed.tex,splncs.bst}
